### 2024-02-02 products/manuscript/Manuscript.qmd
Update Manuscript.qmd

---
title: "Factors Influencing Abundance & Virus Isolation in Mosquitoes"
subtitle: "Which environmental, ecological, and temporal factors impact mosquito populations  "
author: Andreas Handel
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../assets/dataanalysis-references.bib
csl: ../assets/apa.csl
---

The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

# Summary/Abstract
_Write a summary of your project._

{{< pagebreak >}}

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("products","assets","antigen-recognition.png"))
```

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

## Statistical analysis
_Explain anything related to your statistical analyses._

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","height-weight-stratified.png"))
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables-files","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

{{< pagebreak >}}

# References

### 2024-02-02 products/manuscript/Manuscript.qmd
first commit

@@
 ---
-title: "Factors Influencing Abundance & Virus Isolation in Mosquitoes"
+title: "Class Exercise: Factors Influencing Abundance & Virus Isolation in Mosquitoes"
 subtitle: "Which environmental, ecological, and temporal factors impact mosquito populations  "
-author: Andreas Handel
+author: Andrew Ruiz
 date: "`r Sys.Date()`"
@@
 csl: ../assets/apa.csl
+editor: 
+  markdown: 
+    wrap: 72
 ---
-
-The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
-You don't need to have exactly these sections, but the content covering those sections should be addressed.
-
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
 
@@
 library(knitr)
+library(readxl)
+
+here::set_here()
+
 ```
@@
 # Summary/Abstract
-_Write a summary of your project._
 
-{{< pagebreak >}}
+Eastern equine encephalitis (EEE) is caused by an Alphavirus transmitted
+to humans by the bite of an infected mosquito. Human infections are
+rare, but serious. About 30-50% of symptomatic cases lead to death and
+those who survive are left with life-long disability. The transmission
+cycle for EEE includes two distinct cycles: an amplifying enzootic cycle
+in which the virus is transmitted between the Culiseta melanura mosquito
+and birds and an epizootic cycle in which bridge vector mosquitoes
+transmit the EEE virus (EEEv) from birds to humans and other mammals.
+Because C. melanura feeds almost exclusively on birds, it is not
+considered a direct human threat. However, EEEv isolations in C.
+melanura are an early warning sign that EEEv is circulating in the
+ecosystem. Isolations in bridge vector mosquito species indicate
+heightened transmission risk to humans.
+
+This project will lay the foundation for my dissertation research. **The
+goal for this phase of the project is to answer which environmental,
+ecological, and temporal factors influence the transmission cycles for
+EEE in southeastern Massachusetts.**
 
@@
 ## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
+
+Humans and other mammals are dead-end hosts for EEE. About 94% of human
+EEE infections are asymptomatic. While symptomatic cases are rare, the
+consequences can be severe. Mortality rates associated with symptomatic
+infections range from 30 to 50%. Those who survive often experience
+lifelong disability that includes debilitating neurological damage. On
+average, there are 11 cases reported in the US annually. Most occur in
+Massachusetts and Florida. Until recently, nearly all human EEE cases in
+Massachusetts occurred in two counties: Bristol and Plymouth.
+
+Since 2000, there have been 45 human cases recorded in Massachusetts,
+resulting in 22 deaths. There are no human vaccines for EEE, and
+treatment consists of palliative care only. Prevention measures include
+personal behaviors to avoid mosquito bites and decreasing the mosquito
+population through pesticide use and environmental modifications like
+removing standing water. The Bristol County Mosquito Control Project
+(BCMCP) coordinates mosquito surveillance and testing in the county from
+June to October. BCMCP has used the same sentinel collection sites and
+trapping methods for over 20 years. Once trapped, mosquitoes are sorted
+by species, and vector species are submitted to the Massachusetts
+Department of Health’s (MDPH) State Lab for PCR virus testing. Results
+are available within 24 hours of submission. When rates of EEE mosquito
+infections are above a defined threshold, MDPH notifies the local boards
+of health and recommends preventive measures. Recommended measures
+include outreach and education to increase personal protective practices
+(Heymann, 2015). Additionally, MDPH may recommend that cities and towns
+in the affected areas cancel evening outdoor events and discourage
+outdoor activities when mosquito vectors are most active (dawn and dusk)
+(Massachusetts Department of Public Health, n.d.). When infection rates
+are high enough to indicate imminent human transmission, the
+Commonwealth of Massachusetts will recommend and fund aerial pesticide
+applications over the affected areas.
 
 ## Description of data and data source
-_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
+
+*The cornerstone of this research will be the mosquito collection data
+provided by a mosquito control project in SE Massachusetts. A formal
+data-sharing agreement will be in place by January 2024. The dataset
+encompasses nearly two decades of detailed mosquito surveillance. The
+datasets include spatial identifiers. Mosquito collection data will be
+integrated with other datasets using spatial or temporal attributes.
+These datasets will include biological and ecological data,
+environmental and spatial data, and demographic data, which are crucial
+in understanding the transmission dynamics of EEE. Potential datasets
+with their sources are outlined in the table below. Other datasets may
+be included based on availability and need.*
+
+| Data                         | Description                                                                                                                                           | Source                                |
+|------------------|-------------------------------|-----------------------|
+| Mosquito Surveillance Data   | Records of mosquito species counts, locations, and dates of collection.
+Data on the presence of EEE in mosquito populations is particularly valuable.
+| XXX County Mosquito Control           |
+| Weather Data                 | Temperature, humidity, rainfall, and wind speed influence mosquito activity, population dynamics, and virus transmission.
+| NOAA                                  |
+| Bird Population Data         | Since birds are a natural reservoir for EEE, information on bird populations and migration patterns could be relevant.
+| Crowd-sourced bird data               |
+| Human Case Reports           | Data on confirmed human cases of EEE, including location, date of onset, and clinical outcomes.
+| MA Dept of Public Health publications |
+| Veterinary Surveillance Data | Since horses and other mammals can also be affected by EEE, veterinary records could provide early warning signs of virus activity in an area.
+| MA Dept of Public Health              |
+
+### Currently, I have collected 2 datasets that are described below:
+
+#### The first dataset is the mosquito collection dataset
+
+```{r}
+dp1 = here("data", "raw-data", "MOSQ_MADA.xlsx")
+
+mos_data = read_excel(dp1)
+
+head(mos_data)
+summary(mos_data)
+
+```
+
+#### The second dataset is the weather data spanning the same time period as the mosquito collection data
+
+```{r}
+dp2 = here("data", "raw-data", "weather_airport.csv")
+weather = read.csv(dp2, fileEncoding = "UTF-8")
+head(weather)
+summary(weather)
+```
 
 ## Questions/Hypotheses to be addressed
-_State the research questions you plan to answer with this analysis._
 
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key.
+\*This project will lay the foundation for my dissertation research.
+**The goal for this phase of the project is to answer which
+environmental, ecological, and temporal factors influence the
+transmission cycles for EEE in southeastern Massachusetts.** While the
+following questions will not be addressed in this phase, this project
+will hopefully inform research that will answer the following questions:
+
+1\. Can an AI-driven model that leverages historic mosquito
+surveillance, ecological, and environmental data accurately quantify the
+risk of human EEEv infection in southeastern Massachusetts?
+
+2\. What is the potential of machine learning algorithms to identify
+early warning signals for EEE outbreaks, enabling timely public health
+interventions in Massachusetts to prevent human infections?
+
+3\. How can machine learning models use real-time data on mosquito
+activity, environmental factors, and weather to accurately predict when
+the risk of EEE transmission to humans is low enough to lift evening
+outdoor activity restrictions? \*
+
+To cite other work (important everywhere, but likely happens first in
+introduction), make sure your references are in the bibtex file
+specified in the YAML header above (here
+`dataanalysis_template_references.bib`) and have the right bibtex key.
 Then you can include like this:
 
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
+Examples of reproducible research projects can for instance be found in
+[@mckay2020; @mckay2020a].
 
@@
 
-_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
+*Specific methods will be determined when all data sources are
+confirmed.*
 
-## Schematic of workflow
+## The process will involve creating a relational database in which all datasets are joined using spatial or temporal attributes.
+This will lead to a thorough examination of the datasets to understand the distributions, identify missing data and outliers, and uncover potential correlations between variables.
+The EDA will involve using visualizations, geographic information systems, and statistical tests to understand how biological, ecological, environmental, and demographic forces influence the EEE transmission cycles.
+The EDA will inform the preprocessing steps, feature engineering, and selection of appropriate AI models.
 
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
+### Schematic of workflow
 
-```{r}
-#| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
-#| echo: FALSE
-knitr::include_graphics(here("products","assets","antigen-recognition.png"))
-```
+*This section will be updated when all data sources are confirmed*
 
 ## Data aquisition
-_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
+
+With the exception of the mosquito collection data, all other datasets
+are publicly available for use without restrictions.
 
 ## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
+
+*This step will be completed once all data sources are confirmed*
 
 ## Statistical analysis
-_Explain anything related to your statistical analyses._
+
+This exercise will serve as the beginning of the exploratory data
+analysis phase. One of the main objectives for of this phase is describe
+the data, understand its structure, clean data, and identify
+associations within the datasets.
 
@@
 
-_Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project._
-
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-
-```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
-#| echo: FALSE
-resulttable=readRDS("../../results/summarytable.rds")
-knitr::kable(resulttable)
-```
+*These will be generated in later phases of the exercise*
 
@@
 
-_To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
-
-```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
-#| echo: FALSE
-knitr::include_graphics(here("results","height-weight-stratified.png"))
-```
-
 ## Full analysis
-
-_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here._
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
-
-```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-resulttable2 = readRDS(here("results","tables-files","resulttable2.rds"))
-knitr::kable(resulttable2)
-```
 
@@
 
-## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
-
 ## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
 
 ## Conclusions
-_What are the main take-home messages?_
 
-_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_
+*Include citations in your Rmd file using bibtex, the list of references
+will automatically be placed at the end*
 
@@
 
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
+These papers [@mckay2020; @mckay2020a] are good examples of papers
+published using a fully reproducible setup similar to the one shown in
+this template.
 
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.
+Note that this cited reference will show up at the end of the document,
+the reference formatting is determined by the CSL file specified in the
+YAML header. Many more style files for almost any journal [are
+available](https://www.zotero.org/styles). You also specify the location
+of your bibtex reference file in the YAML. You can call your reference
+file anything you like, I just used the generic word `references.bib`
+but giving it a more descriptive name is probably better.
 

### 2024-02-02 products/manuscript/Manuscript.qmd
Final push before submission
Updated qmd. removed bib and csl path for now.

@@
     highlight-style: github
-bibliography: ../assets/dataanalysis-references.bib
-csl: ../assets/apa.csl
 editor: 
@@
 
-{{< pagebreak >}}
-
 # Results
@@
 ## Full analysis
-
-{{< pagebreak >}}
 

### 2024-02-02 products/manuscript/Manuscript.qmd
final push

@@
 head(mos_data)
-summary(mos_data)
 
@@
 head(weather)
-summary(weather)
 ```

### 2024-02-23 R/processing-code/MADAproject_pt2_ruiz.qmd
Part 2
This commit includes the work associated with part 2 of the project including the processing code located here: /R/processing-code/MADAproject_pt2_ruiz.qmd

---
title: "MADA Project Part 2"
subtitle: "loading, cleaning, exploring the data"
author: "Andrew Ruiz"
date: 2024-02-23
format: html
editor: visual
---

# Introduction
## This file contains the process for loading, cleaning and exploring the datasets.

# Loading libraries
```{r, message=FALSE, warning=FALSE}
library(readxl) #for loading Excel files
library(httr)   #for getting data from the web
library(dplyr)  #for data processing/cleaning
library(tidyr)  #for data processing/cleaning
library(skimr)  #for visualizing data 
library(here)   #allows relative pathways for data
library(readr)
library(naniar) 
library(ggplot2)
library(sf)
library(topicmodels)
library(openxlsx)
library(lubridate)
library(knitr)
library(kableExtra)
library(webshot2)
library(magick)
library(leaflet)
library(htmlwidgets)
```
# Loading data

## Mosquito trap data
### This dataset contains trap counts by species for all mosquito traps set in the Bristol County from 2007-2023
```{r trap-data-import}
# Construct the path to the Excel file
file_path_mos <- here("data", "processed-data", "MOSQ_MADA.xlsx")

# import the data to R and store it in a dataframe named "mosquito"
mosquito_raw <- read_excel(file_path_mos)

#load the first few rows of data to verify the data loaded as expected
head(mosquito_raw)
```
## Mosquito trap sites 
### Location for the mosquito trap sites
```{r}
# Construct the path to the file
file_path_trap = here("data", "processed-data", "BCMCP_trap_sites_2021.xlsx")

# import the data to R and store it in a dataframe named "trap_location"
trap_location = read_excel(file_path_trap)

#load the first few rows of data to verify the data loaded as expected
head(trap_location)
str(trap_location)

#The last 3 columns have missing data. However, they are not important for the purposes of this project
# there is not a need to delete the records
```
## Mosquito virus data
### The virus isolation data is stored in separate spreadsheets by year from 2014 to 2020
### This code will load all the files, combine them, and then save the combined file in the "raw data" folder
```{r}
#Define the path to the folder with Excel files
folder_path <- here("data", "processed-data", "virus_data")

#List all Excel files (.xls and .xlsx)
file_paths <- list.files(path = folder_path, pattern = "\\.xls[x]*$", full.names = TRUE)

#Read each file into a list of data frames
data_list <- lapply(file_paths, read_excel)

#Check the names of the files read
print(basename(file_paths))

#Combine the datasets
combined_data <- bind_rows(data_list)

# Define the path for the combined dataset
combined_path <- here("data", "processed-data", "combined_virus_data.csv")

# Save the combined data frame as a CSV file
write.csv(combined_data, combined_path, row.names = FALSE)

# View the first few rows to confirm it's loaded correctly
head(virus_data)
```
## Bird data
### The complete bird dataset is over 600 MB. Because it is too large for GitHub, it is stored locally. It contain variables and records that are not needed for this analysis. Eight of 50 varibles will be kept and the records will be filtered to show only birds from the Genus Corvus observed in Massachusetts after 2014-01-01
```{r}
# Load complete bird data
bird <- read.table("/Users/andrewruiz/0011508-240216155721649.csv", sep = "\t", header = TRUE, fill = TRUE)

# Select variables to keep and then filter records based on order, stateProvince, and eventDate
bird_select <- bird %>%
  select(genus, 
         species,
         stateProvince, 
         decimalLatitude, 
         decimalLongitude, 
         eventDate) %>%
  mutate(eventDate = as.Date(eventDate, format = "%Y-%m-%d")) %>%
  filter(eventDate > as.Date("2014-01-01"), genus == "Corvus", stateProvince == "Massachusetts")

# Inspect the structure of the filtered dataset
str(bird_select)

# Define the file path using here()
output_file_path <- here("data", "processed-data", "filtered_bird_data.csv")

# Save the bird_select dataframe to the specified folder
write.csv(bird_select, output_file_path, row.names = FALSE)

# Load the filtered bird data to R
filtered_bird = read_csv(output_file_path, show_col_types = FALSE)

# ensure the csv loaded properly
str(filtered_bird)
head(filtered_bird)
```

## Weather data
### This dataset contains daily weather summaries from the Taunton, MA Airport from 2007-2023
```{r}
# Construct the path to the file
file_path_wx = here("data", "processed-data", "weather_airport.csv")

# import the data to R and store it in a dataframe named "wx"
wx = read.csv(file_path_wx)

#load the first few rows of data to verify the data loaded as expected
head(wx)
```
## Human and veterinary cases of eastern equine encephalitis and West Nile virus
```{r}
# Construct the path to the human case data file
file_path_human <- here("data", "processed-data", "human_cases.xlsx")

# import the data to R and store it in a dataframe named "cases_human"
cases_human <- read_excel(file_path_human)

#load the first few rows of data to verify the data loaded as expected
head(cases_human)

# Construct the path to the veterinary case data file
file_path_vet <- here("data", "processed-data", "animal_cases.xlsx")

# import the data to R and store it in a dataframe named "cases_human"
cases_vet <- read_excel(file_path_vet)

#load the first few rows of data to verify the data loaded as expected
head(cases_vet)
```

# Processing and cleaning the Mosquito trap data
### Examine the variables and change all column names to lower case
```{r}
str(mosquito_raw)

# Set column names to lowercase
names(mosquito_raw) <- tolower(names(mosquito_raw))

# Check the first few rows to confirm changes
head(mosquito_raw)

```

### Identify the time period the dataset covers
```{r}
earliest_date <- min(mosquito_raw$`Collection Date`, na.rm = TRUE)
latest_date <- max(mosquito_raw$`Collection Date`, na.rm = TRUE)
print(paste("Date range:", earliest_date," to ", latest_date))
```
#### The mosquito dataset spans 2007 to 2023. However, the virus and case data only covers the years 2014 to 2020. Because mosquito populations can be influences by processes of the previous year. We will set the analytical horizon will span 2013 to 2021.*Collection data before 2013 and after 2021 will be removed* along with records with important missing variables identified below.

### Identify missing data
```{r}
sapply(mosquito_raw, function(x) sum(is.na(x)))
```
#### The majority of the missing data is in the "Submitted for Testing" variable. This is not a crucial variable, so we can keep those records.The other variable are important and so we will remove those records with the other missing data.

### The Massachussetts Department of Public Health adpoted a taxonomy change before it was fully vetted. The change divided the genus *Aedes* into a new genus: *Ochlerotatus*. To align with accepted standards, I will change them back to *Aedes*.
```{r}
# Get unique values using dplyr
unique_genus_values <- mosquito_raw %>%
  distinct(genus) %>%
  pull(genus)

# Print the unique genus values
print(unique_genus_values)

# Change records where genus is "Ochlerotatus" to "Aedes"
mosquito_raw$genus[mosquito_raw$genus == "Ochlerotatus"] <- "Aedes"

# Check the first few rows to confirm changes, or use unique() to see the change in genus values
unique(mosquito_raw$genus)

```

### Execute proposed changes.
```{r}
# Create a copy of the dataframe to avoid modifying the original
mosquito_modified <- mosquito_raw

# Change records where genus is "Ochlerotatus" to "Aedes"
#mosquito_modified$genus[mosquito_modified$genus == "Ochlerotatus"] <- "Aedes"

# Filter out records where Collection Date is before 2013 or after 2021
# Ensure Collection Date is in Date format if it's not already
mosquito_modified$`collection date` <- as.Date(mosquito_modified$`collection date`)

# Convert all species code values to uppercase
mosquito_modified <- mosquito_modified %>%
  mutate(`species code` = toupper(`species code`))

# Now apply the date filter
mosquito_modified <- mosquito_modified[mosquito_modified$`collection date` >= as.Date("2013-01-01") & 
                                       mosquito_modified$`collection date` <= as.Date("2021-12-31"), ]

# Remove rows with NA values in the remaining columns of the modified dataframe, 
# except for "Submitted for Testing" which we'll handle separately
temp_columns <- names(mosquito_modified)
temp_columns <- temp_columns[temp_columns != "submitted for testing"]

# Applying NA omit on selected columns only
mosquito_clean_except_submitted <- na.omit(mosquito_modified[temp_columns])

# Since na.omit on selected columns messes with row alignment, let's filter the original dataset based on the cleaned IDs
clean_ids <- row.names(mosquito_clean_except_submitted)
mosquito_final <- mosquito_modified[clean_ids, ]

# Add back "Submitted for Testing" from the original dataframe if needed
# Note: This step might be redundant if "Submitted for Testing" was not actually removed or if it's included in the cleaning process
mosquito_final$`submitted for testing` <- mosquito_raw$`submitted for testing`[clean_ids]

# Check the structure of the final cleaned dataframe
str(mosquito_final)
```

## Reformat the collection data
### The Massachusetts Department of Public Health requires that all mosquito control district submit their colelction data in a specific format. This format divides each trap event into separate rows based on mosquito species. However, this format is not useful for certain calculations where a zero count for a species in not explicitly recorded. Even if you sort by species and calculate average count per trap, the average could be inflated since the records will not include trap events with a zero count of that species. 
### In order to correct this, I will pivot the data and create a column for every unique "species code" and sum the "pool size" for each species code. The resulting table will have one row for every trap event -a trap event is when "town", "date of collection", "trap type" are all the same.

```{r}
library(dplyr)
library(tidyr)

# Group and summarize the mosquito data to prepare for wide format transformation.
# This step aggregates the total 'pool size' for each unique combination of 
# 'collection date', 'town', 'trap type', 'species code', and 'mcd' (Mosquito Control District),
# while excluding 'submitted for testing' from the grouping to avoid potential inconsistencies.
mosquito_summarized <- mosquito_final %>%
  group_by(`collection date`, town, `trap type`, `species code`, mcd, `submitted for testing`) %>%
  summarise(pool_size_sum = sum(`pool size`, na.rm = TRUE), .groups = 'drop')
   # Ensure that the summarization does not drop any groups by setting .groups = 'drop'.

# Transform the summarized mosquito data to a wide format where each 'species code' becomes a separate column.
# This pivot operation facilitates easier analysis of mosquito counts by species across different events.
# 'values_fill' is used to fill in zeroes for any 'species code' not caught in a specific trapping event,
# ensuring that the dataset has a consistent shape and making it more analyzable.
wide_data <- mosquito_summarized %>%
  pivot_wider(
    names_from = `species code`,
    values_from = pool_size_sum,
    values_fill = list(pool_size_sum = 0),
    id_cols = c(`collection date`, town, `trap type`, mcd)
  )

# Output the transformed wide-format data to inspect its structure and content.
# This step is crucial for verifying that the pivot operation produced the expected wide-format dataset,
# with each row representing a unique trapping event and columns for each mosquito species.
head(wide_data)

# Save the new file
# Specify the file path using here()
file_path_wide <- here("data", "processed-data", "mosquito_wide.csv")

# Save the wide-format data to the specified path
write_csv(wide_data, file_path_wide)

# Print the file path to confirm where the file has been saved
print(paste("File saved to:", file_path_wide))
```

# Exploring the Data
##Study location
Below is a map of the mosquito trap sites. This is an interactive map and will only work on web or other html-friendly platforms. 
```{r}
# Create the map
map_study <- leaflet(trap_location) %>%
  addTiles() %>%  # This adds the default OpenStreetMap tiles
  addMarkers(lng = ~DecLong, lat = ~DecLat, popup = ~Location)  # Customize the popup content as needed

# Display the map
map_study

# Specify the path and name of your HTML file
file_path_map <- here("results", "figures", "map_study.html")

# Save the map
saveWidget(map_study, file_path_map, selfcontained = TRUE)
```

## EEE in mosquitoes and mammals
### To give an idea of when EEE virus silled over from mosquitoes and birds and into mammal populations, lets take a look at the data. Let's see when the first EEE isolations were found in mosquitoes each year. Let's also look at the years that human and veterinary cases occured.
### From the table, we see that human or othe mammal cases were recorded in 2014, 2018, 2019, and 2020. 2019 had the highest case counts for both humans and other mammals. **Because of this, we will pay close attention to 2019 through out the exploration.**

```{r}
library(dplyr)
library(lubridate)

# virus_data, cases_human, and cases_vet data frames

# Prepare virus_data: Find the first positive EEE test date by year
first_positive_eee_test_by_year <- virus_data %>%
  filter(Virus == "EEE") %>%
  mutate(Year = year(as.Date(Collection.Date))) %>%
  group_by(Year) %>%
  summarise(First_Positive_Test_Date = min(Collection.Date)) %>%
  ungroup() %>%
  arrange(Year)

# Prepare cases_human: Summarize human EEE cases by year
human_cases_by_year <- cases_human %>%
  filter(`Virus Result` == "EEE") %>%
  mutate(Year = year(`Onset Date`)) %>%
  group_by(Year) %>%
  summarise(Human_Cases = n()) %>%
  ungroup() %>%
  arrange(Year)

# Prepare cases_vet: Summarize animal EEE cases by year
animal_cases_by_year <- cases_vet %>%
  filter(virus == "eee") %>%
  group_by(year) %>%
  summarise(Animal_Cases = n()) %>%
  ungroup() %>%
  arrange(year)

# Ensure 'Year' column is numeric across all data frames for compatibility
# Note: This step might already be covered by the mutations above but included here for clarity
first_positive_eee_test_by_year$Year <- as.numeric(first_positive_eee_test_by_year$Year)
animal_cases_by_year$year <- as.numeric(animal_cases_by_year$year) # Make sure year is numeric and named consistently

# Merge the summaries into one combined data frame
combined_eee_data <- first_positive_eee_test_by_year %>%
  left_join(human_cases_by_year, by = "Year") %>%
  left_join(animal_cases_by_year, by = c("Year" = "year")) # Ensure correct column names are used for joining

# Replace NA values with 0 for Human_Cases and Animal_Cases
combined_eee_data$Human_Cases[is.na(combined_eee_data$Human_Cases)] <- 0
combined_eee_data$Animal_Cases[is.na(combined_eee_data$Animal_Cases)] <- 0

# Use kable() to create a basic table
library(knitr)
library(kableExtra)

# Creating the table with centered headings and data
kable_table_isolation_case <- kable(combined_eee_data, "html", 
                     col.names = c("Year", "First Positive EEE Test Date", "Human EEE Cases", "Animal EEE Cases"),
                     caption = "Annual Summary of EEE Virus Activity",
                     align = c('c','c','c','c')) %>% # This aligns all columns' data to center
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = F, position = "center") %>%
  add_header_above(c(" " = 1, "EEE Virus Detection and Cases" = 3)) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, width = "8em") %>% # Adjusted width without altering alignment
  column_spec(3, width = "5em", bold = TRUE, color = "red") %>%
  column_spec(4, width = "5em", bold = TRUE, color = "blue")

# Displaying the table
kable_table_isolation_case

# Define the path where you want to save the HTML file
#output_path_quarto <- here("results", "tables", "eee_isolation_cases")
output_path_png <- here("results", "tables", "eee_isolation_cases.png")

# Save the kableExtra table as an HTML file
#save_kable(kable_table_isolation_case, file = output_path_quarto)
save_kable(kable_table_isolation_case, file = output_path_png)
```
## Which county has the most EEE positive mosquitoes of those tested? 
```{r}
eee_positive_by_county <- virus_data %>%
  filter(Virus == "EEE") %>%  # Filter for rows where Virus is EEE
  group_by(County) %>%
  summarise(EEE_Positive_Tests = n()) %>%  # Count the number of EEE positive tests in each group
  arrange(desc(EEE_Positive_Tests))  # Order the results by EEE_Positive_Tests in descending order

# Create the table with kable and style it with kableExtra
eee_positive_table <- kable(eee_positive_by_county, "html", 
                            col.names = c("County", "EEE Positive Tests"),
                            caption = "EEE Positive Tests by County",
                            align = c('c','c')) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = FALSE, position = "center") %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, bold = TRUE, color = "red")

# Display the table
eee_positive_table

# Define the path where you want to save the HTML file
#output_path_qmd <- here("results", "tables", "eee_isolation_county.html")
output_path_county <- here("results", "tables", "eee_isolation_county.png")

# Save the kableExtra table as an HTML file
#save_kable(eee_positive_table, file = output_path_qmd)
save_kable(eee_positive_table, file = output_path_county)
```
## *Culiseta melanura* (MEL) is the main enzootic vector of EEE. Some entomologist believe that high levels of MEL traps indicate higher risk for EEE transmission. Using the new mosquito collection table created from the pivot, let's calculate the proportion of MEL captured compared to all other mosquito vector species and create a plot to visualize it. 
### 
```{r}
# Define path to the data file
trap_wide_path = here("data", "processed-data", "mosquito_wide.csv")

# Read the data
trap_proportion = read.csv(trap_wide_path)

# Convert 'collection.date' to Date type using the correct format and extract the year
trap_proportion$Year <- year(ymd(trap_proportion$collection.date))

# Sum all specified species by the calculated year, including any additional species codes you have
yearly_sums <- trap_proportion %>%
  group_by(Year) %>%
  summarise(across(c("MEL", "PER", "PIP", "RES", "SAL", "TRI", "SAP"), sum, na.rm = TRUE)) %>%
  # After summarising, calculate the total counts for all species per year
  mutate(Total = rowSums(across(c("MEL", "PER", "PIP", "RES", "SAL", "TRI", "SAP"))),
         # Calculate the proportion of MEL out of the total for all species
         MEL_Prop = MEL / Total)

# View the result with the proportion of MEL
print(yearly_sums)

library(ggplot2)

mel_plot = ggplot(yearly_sums, aes(x = factor(Year), y = MEL_Prop, fill = "MEL")) +
  geom_bar(stat = "identity", color = "black") +
  labs(title = expression(Proportion ~ of ~ italic("Culiseta melanura") ~ (MEL) ~ Mosquitoes ~ by ~ Year),
       x = NULL,
       y = "Proportion of MEL") +
  scale_fill_discrete(name = "Species") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
mel_plot
# Specify the file path where you want to save the plot
#file_path_mel = here("results", "figures", "mel_proportion.png")
# Use ggsave to save the plot
#ggsave(file_path_mel, plot = mel_plot, width = 8, height = 6, dpi = 300)

```
## Mosquito abundance is affected by weather. Mosquitoes depend on standing water to complete their larval development. Let's take a look at the annual precipitation using the wx data frame. 
```{r annual precip}

# Convert DATE from character to Date type using as.Date and specifying the format
wx$DATE <- as.Date(wx$DATE, format="%m/%d/%y")

# The as.Date function treats years in the format "yy" as follows:
# Years 00-68 are treated as 2000-2068, and years 69-99 are treated as 1969-1999.
# Adjust this according to your data's actual year range if needed.

# Extract the year from each DATE
wx$Year <- format(wx$DATE, "%Y") %>% as.numeric()

# Now that we have the years as numeric values, let's filter for 2013 to 2020
filtered_wx <- wx %>%
  filter(Year >= 2013 & Year <= 2020)

# Sum PRCP by Year for the filtered data
summed_prcp_by_year_filtered <- filtered_wx %>%
  group_by(Year) %>%
  summarise(Total_PRCP = sum(PRCP, na.rm = TRUE))

# View the result
print(summed_prcp_by_year_filtered)

# Create the ggplot object and assign it to a variable
precip_plot <- ggplot(summed_prcp_by_year_filtered, aes(x = as.factor(Year), y = Total_PRCP)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Total Precipitation by Year (2013-2020)",
       x = "Year",
       y = "Total Precipitation (inches)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Specify the file path where you want to save the plot
file_path_precip <- here("results", "figures", "annual_precip.png")

# Use ggsave to save the plot, make sure to use the plot object
ggsave(file_path_precip, plot = precip_plot, width = 8, height = 6, dpi = 300)
```

### 2024-02-23 products/manuscript/Manuscript.qmd
Part 2
This commit includes the work associated with part 2 of the project including the processing code located here: /R/processing-code/MADAproject_pt2_ruiz.qmd

@@
 ---
-title: "Class Exercise: Factors Influencing Abundance & Virus Isolation in Mosquitoes"
+title: "DRAFT: Factors Influencing Abundance & Virus Isolation in Mosquitoes"
 subtitle: "Which environmental, ecological, and temporal factors impact mosquito populations  "
@@
 include outreach and education to increase personal protective practices
-(Heymann, 2015). Additionally, MDPH may recommend that cities and towns
-in the affected areas cancel evening outdoor events and discourage
-outdoor activities when mosquito vectors are most active (dawn and dusk)
-(Massachusetts Department of Public Health, n.d.). When infection rates
-are high enough to indicate imminent human transmission, the
+. Additionally, MDPH may recommend that cities and towns in the affected
+areas cancel evening outdoor events and discourage outdoor activities
+when mosquito vectors are most active (dawn and dusk). When infection
+rates are high enough to indicate imminent human transmission, the
 Commonwealth of Massachusetts will recommend and fund aerial pesticide
@@
 | Bird Population Data         | Since birds are a natural reservoir for EEE, information on bird populations and migration patterns could be relevant.
-| Crowd-sourced bird data               |
+| eBird                                 |
 | Human Case Reports           | Data on confirmed human cases of EEE, including location, date of onset, and clinical outcomes.
@@
 | MA Dept of Public Health              |
-
-### Currently, I have collected 2 datasets that are described below:
-
-#### The first dataset is the mosquito collection dataset
-
-```{r}
-dp1 = here("data", "raw-data", "MOSQ_MADA.xlsx")
-
-mos_data = read_excel(dp1)
-
-head(mos_data)
-
-```
-
-#### The second dataset is the weather data spanning the same time period as the mosquito collection data
-
-```{r}
-dp2 = here("data", "raw-data", "weather_airport.csv")
-weather = read.csv(dp2, fileEncoding = "UTF-8")
-head(weather)
-```
 
@@
 
-To cite other work (important everywhere, but likely happens first in
-introduction), make sure your references are in the bibtex file
-specified in the YAML header above (here
-`dataanalysis_template_references.bib`) and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in
-[@mckay2020; @mckay2020a].
-
 {{< pagebreak >}}
@@
 
-*This section will be updated when all data sources are confirmed*
+*This section will be updated before analysis begins*
 
@@
 
-*This step will be completed once all data sources are confirmed*
+Datasets will be loaded into R in a Quarto document to document and
+describe cleaning methods. There are two datasets that will require more
+extensive cleaning: the mosquito trap data and the crowd-sourced bird
+data. The process is outlined in detail in the MADAproject_pt2.qmd file.
+Less significant cleaning and data management steps not included in this
+manuscript are also outlined in that file.
+
+The mosquito dataset spans 2007 to 2023. However, the virus and case
+data only covers the years 2014 to 2020. Because mosquito populations
+can be influences by processes of the previous year. We will set the
+analytical horizon will span 2013 to 2021.
+
+Additionally, the Massachusetts Department of Public Health requires
+that all mosquito control district submit their colelction data in a
+specific format. This format divides each trap event into separate rows
+based on mosquito species. However, this format is not useful for
+certain calculations where a zero count for a species is not explicitly
+recorded. For example, even when sorted by species to calculate the
+average count per trap, the average could be inflated since the records
+will not include trap events with a zero count of that species. In order
+to correct this, the data was pivoted to create a column for every
+unique "species code" then the "pool size" for each species code is
+summed for each trap event. A trap event is when "town", "date of
+collection", "trap type" are all the same.. For species codes not
+included in the orginal table for a trap event, the field will be
+assigned a zero value. The resulting table will have one row for every
+trap event.
+
+The bird observation dataset was obtained from Cornell University's
+eBird program. The original dataset included all reported bird
+observations for Southeast Massachusetts and was over 600 MB of data.
+While all birds are potential amplifying hosts for EEE and West Nile
+Virus (WNV), birds from the Genus *Corvus* are of particular concern.
+The bird dataset processed to include records of *Corvus spp* sightings
+after 2014. The final dataset included 8 of the 50 variables, including
+location data.
 
@@
 
+The first phase of the exploration is complete. The next phase will
+include statistical tests to determine if there are possible
+associations between the environmental variables and the mosquito
+abundance as measured by trap counts or EEE virus activity.
+
+One factor to consider is if early virus isolations are associated with
+cases in humans and other mammals. The table below shows the date of the
+first EEE virus isolation in mosquitoes and the number of human of other
+animal cases by year.
+
+![](images/eee_isolation_cases-01.png){fig-align="center" width="320"}
+
+Unfortunately, the time horizon is limited by data availability.
+Additionally, EEE human infections are rare, so traditional statistical
+methods may be inadequate to suggest any relationships. What is clear is
+that 2019 was an exceptional year for EEE.
+
+*Culiseta melanura* (MEL) plays an important role in EEE virus
+amplification in the enzootic cycle. Understanding the population
+dynamics might provide more insight into what is driving the 2019 spike.
+
+The table below shows the proportion of MEL among all mosquito vector
+species trapped in Bristol County, MA.
+
+![](images/mel_proportion.png){fig-align="center"}
+
+From this we see a large proportion of MEL in the traps for 2019. In
+fact, they comprise more than half of all vector species for that year.
+Another interesting aspect is the high proportion in 2018. MEL
+overwinters as larvae and as long as there is sufficient standing water
+to last the winter, MEL hatched in the previous fall would survive to
+emerge the next year. This suggests it might be important to look at
+conditions in the previous year or season for a better understanding.
+
+Since water plays a significant role in the mosquito life cycle, let's
+look at the annual precipitation.
+
+![](images/annual_precip-02.png){fig-align="center"}
+
+The graph indicates that the annual precipitation total for 2018 is
+higher than the other years. This suggest that there was enough water to
+support large MEL larval populations though out the winter.
+
+## Basic statistical analysis
 *These will be generated in later phases of the exercise*
 
-## Basic statistical analysis
+## Fullanalysis
 
-## Full analysis
-
-# Discussion
+*These will be generated in later phases of the exercise* 
+# Discussion *this section will be completed when analysis is done*
 
 ## Strengths and Limitations 
+*this section will be updated after analysis is complete*
 
-## Conclusions
+While the reliability of the mosquito trap data is high, the trap data
+cannot be directly associated with the virus isolation data. The
+information this scale is not released by the State. Spatial scale is
+also an issue with the human and veterinary cases data. Human case data
+is released only on a county scale to ensure patient privacy. Some
+veterinary case data is available on the town-level, but most remains
+only on the county-level.
 
-*Include citations in your Rmd file using bibtex, the list of references
-will automatically be placed at the end*
+These limitations are barriers to accurate spatial analysis.
 
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers
-published using a fully reproducible setup similar to the one shown in
-this template.
-
-Note that this cited reference will show up at the end of the document,
-the reference formatting is determined by the CSL file specified in the
-YAML header. Many more style files for almost any journal [are
-available](https://www.zotero.org/styles). You also specify the location
-of your bibtex reference file in the YAML. You can call your reference
-file anything you like, I just used the generic word `references.bib`
-but giving it a more descriptive name is probably better.
-
+## Conclusions *this section will be completed when analysis is done*
 {{< pagebreak >}}
@@
 # References
+
+References will be added in part 3 of the project

### 2024-02-23 R/processing-code/MADAproject_pt2_ruiz.qmd
Part 2
corrected typos

@@
 library(here)   #allows relative pathways for data
-library(readr)
-library(naniar) 
-library(ggplot2)
-library(sf)
-library(topicmodels)
-library(openxlsx)
-library(lubridate)
-library(knitr)
-library(kableExtra)
-library(webshot2)
-library(magick)
-library(leaflet)
-library(htmlwidgets)
+library(readr) #reading and writing data to and from CSV, TSV, and fixed-width files
+library(naniar) #tools to work with missing data
+library(ggplot2) #plot creation
+library(sf) #standardized way to encode spatial data
+library(topicmodels) #tools for topic modeling on text data
+library(openxlsx) #Facilitates reading from and writing to .xlsx files
+library(lubridate) #Simplifies the work with dates and times
+library(knitr) #Allows for dynamic report generation in R
+library(kableExtra) #generates simple HTML or LaTeX tables
+library(webshot2) #tools to take screenshots of web pages 
+library(magick) #advanced image processing in R
+library(leaflet) #Enables the creation of interactive maps
+library(htmlwidgets) #Facilitates the creation of R bindings to JavaScript libraries
 ```
@@
 ### This dataset contains trap counts by species for all mosquito traps set in the Bristol County from 2007-2023
+### this dataset will undergo extensive processing below. 
 ```{r trap-data-import}
@@
 head(wx)
+
+#the columns below are the only ones needed for this project. 
+wx_selected <- wx %>%
+  select(STATION, NAME, DATE, PRCP)
+
+head(wx_selected)
 ```
@@
 head(cases_vet)
+#date of onset is missing for some records, however, they are not needed for this project.
+
 ```

### 2024-02-23 R/processing-code/MADAproject_pt2_ruiz.qmd
Part 2 final
Corrected a pathway in the qmd file

@@
 # Introduction
+
 ## This file contains the process for loading, cleaning and exploring the datasets.
@@
 # Loading libraries
+
 ```{r, message=FALSE, warning=FALSE}
@@
 ```
+
 # Loading data
@@
 ## Mosquito trap data
+
 ### This dataset contains trap counts by species for all mosquito traps set in the Bristol County from 2007-2023
+
 ### this dataset will undergo extensive processing below.
+
 ```{r trap-data-import}
@@
 ```
+
 ## Mosquito trap sites
+
 ### Location for the mosquito trap sites
+
 ```{r}
@@
 # there is not a need to delete the records
+#However, we will delete the columns
+
+trap_location_selected <- trap_location %>%
+  select(`Trap Site Address`, MCD, Location, DecLat, DecLong)
+trap_location_selected
 ```
+
 ## Mosquito virus data
+
 ### The virus isolation data is stored in separate spreadsheets by year from 2014 to 2020
+
 ### This code will load all the files, combine them, and then save the combined file in the "raw data" folder
+
 ```{r}
@@
 ```
+
 ## Bird data
+
 ### The complete bird dataset is over 600 MB.
@@
 Eight of 50 varibles will be kept and the records will be filtered to show only birds from the Genus Corvus observed in Massachusetts after 2014-01-01
+
 ```{r}
@@
 ## Weather data
+
 ### This dataset contains daily weather summaries from the Taunton, MA Airport from 2007-2023
+
 ```{r}
@@
 ```
+
 ## Human and veterinary cases of eastern equine encephalitis and West Nile virus
+
 ```{r}
@@
 # Processing and cleaning the Mosquito trap data
+
 ### Examine the variables and change all column names to lower case
+
 ```{r}
@@
 ### Identify the time period the dataset covers
+
 ```{r}
@@
 ```
+
 #### The mosquito dataset spans 2007 to 2023.
@@
 ### Identify missing data
+
 ```{r}
@@
 ```
+
 #### The majority of the missing data is in the "Submitted for Testing" variable.
@@
 To align with accepted standards, I will change them back to *Aedes*.
+
 ```{r}
@@
 ### Execute proposed changes.
+
 ```{r}
@@
 ## Reformat the collection data
+
 ### The Massachusetts Department of Public Health requires that all mosquito control district submit their colelction data in a specific format.
@@
 Even if you sort by species and calculate average count per trap, the average could be inflated since the records will not include trap events with a zero count of that species.
+
 ### In order to correct this, I will pivot the data and create a column for every unique "species code" and sum the "pool size" for each species code.
@@
 # Exploring the Data
-##Study location
-Below is a map of the mosquito trap sites. This is an interactive map and will only work on web or other html-friendly platforms. 
+
+##Study location Below is a map of the mosquito trap sites. This is an interactive map and will only work on web or other html-friendly platforms.
+
 ```{r}
@@
 ## EEE in mosquitoes and mammals
+
 ### To give an idea of when EEE virus silled over from mosquitoes and birds and into mammal populations, lets take a look at the data.
@@
 Let's also look at the years that human and veterinary cases occured.
+
 ### From the table, we see that human or othe mammal cases were recorded in 2014, 2018, 2019, and 2020.
@@
 ```
+
 ## Which county has the most EEE positive mosquitoes of those tested?
+
 ```{r}
@@
 ```
+
 ## *Culiseta melanura* (MEL) is the main enzootic vector of EEE.
@@
 Using the new mosquito collection table created from the pivot, let's calculate the proportion of MEL captured compared to all other mosquito vector species and create a plot to visualize it.
+
 ###
+
 ```{r}
@@
 ```
+
 ## Mosquito abundance is affected by weather. Mosquitoes depend on standing water to complete their larval development. Let's take a look at the annual precipitation using the wx data frame.
+
 ```{r annual precip}

### 2024-02-23 products/manuscript/Manuscript.qmd
Part 2 final
Corrected a pathway in the qmd file

@@
 
+It is important to know where the mosquitoes infected with EEE are captured. Below is a table showing the virus isolations for EEE in trapped mosquitoes by county between 2014 and 2020. 
+```{r}
+# Create the table with kable and style it with kableExtra
+eee_positive_table <- kable(eee_positive_by_county, "html", 
+                            col.names = c("County", "EEE Positive Tests"),
+                            caption = "EEE Positive Tests by County",
+                            align = c('c','c')) %>%
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
+                full_width = FALSE, position = "center") %>%
+  column_spec(1, bold = TRUE) %>%
+  column_spec(2, bold = TRUE, color = "red")
+
+# Display the table
+eee_positive_table
+```
+The geographic focus for this study is Bristol County.
 One factor to consider is if early virus isolations are associated with
@@
 animal cases by year.
+```{r}
+# Use kable() to create a basic table
+library(knitr)
+library(kableExtra)
 
-![](images/eee_isolation_cases-01.png){fig-align="center" width="320"}
+# Creating the table with centered headings and data
+kable_table_isolation_case <- kable(combined_eee_data, "html", 
+                     col.names = c("Year", "First Positive EEE Test Date", "Human EEE Cases", "Animal EEE Cases"),
+                     caption = "Annual Summary of EEE Virus Activity",
+                     align = c('c','c','c','c')) %>% # This aligns all columns' data to center
+  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
+                full_width = F, position = "center") %>%
+  add_header_above(c(" " = 1, "EEE Virus Detection and Cases" = 3)) %>%
+  column_spec(1, bold = TRUE) %>%
+  column_spec(2, width = "8em") %>% # Adjusted width without altering alignment
+  column_spec(3, width = "5em", bold = TRUE, color = "red") %>%
+  column_spec(4, width = "5em", bold = TRUE, color = "blue")
+
+# Displaying the table
+kable_table_isolation_case
+```
 
@@
 
-The table below shows the proportion of MEL among all mosquito vector
+The graph below shows the proportion of MEL among all mosquito vector
 species trapped in Bristol County, MA.

### 2024-03-04 R/processing-code/MADAproject_pt2_ruiz.qmd
Part 2 Render correction
located source of error preventing the qmd from rendering.

@@
 #Combine the datasets
-combined_data <- bind_rows(data_list)
+combined_virus_data <- bind_rows(data_list)
 
@@
 # Save the combined data frame as a CSV file
-write.csv(combined_data, combined_path, row.names = FALSE)
+write.csv(combined_virus_data, combined_path, row.names = FALSE)
 
 # View the first few rows to confirm it's loaded correctly
-head(virus_data)
+head(combined_virus_data)
 ```
@@
 #date of onset is missing for some records, however, they are not needed for this project. 
-
 ```
@@
 ```{r}
-earliest_date <- min(mosquito_raw$`Collection Date`, na.rm = TRUE)
-latest_date <- max(mosquito_raw$`Collection Date`, na.rm = TRUE)
+earliest_date <- min(mosquito_raw$`collection date`, na.rm = TRUE)
+latest_date <- max(mosquito_raw$`collection date`, na.rm = TRUE)
 print(paste("Date range:", earliest_date," to ", latest_date))
@@
 # Prepare virus_data: Find the first positive EEE test date by year
-first_positive_eee_test_by_year <- virus_data %>%
+first_positive_eee_test_by_year <- combined_virus_data %>%
   filter(Virus == "EEE") %>%
-  mutate(Year = year(as.Date(Collection.Date))) %>%
+  mutate(Year = year(as.Date(`Collection Date`))) %>%
   group_by(Year) %>%
-  summarise(First_Positive_Test_Date = min(Collection.Date)) %>%
+  summarise(First_Positive_Test_Date = min(`Collection Date`)) %>%
   ungroup() %>%
@@
 ```{r}
-eee_positive_by_county <- virus_data %>%
+eee_positive_by_county <- combined_virus_data %>%
   filter(Virus == "EEE") %>%  # Filter for rows where Virus is EEE

### 2024-03-07 R/processing-code/MADAproject_pt2_ruiz.qmd
Fixed manuscript render error
loaded package and inserted image generated in processing qmd

@@
 # Construct the path to the Excel file
-file_path_mos <- here("data", "processed-data", "MOSQ_MADA.xlsx")
+#file_path_mos <- here("data", "raw-data", "MOSQ_MADA.xlsx")
+file_path_mos <- here("data", "raw-data", "MOSQ_MADA.xlsx")
+file_exists_check <- file.exists(file_path_mos)
+file_path_mos <- here("data", "raw-data", "MOSQ_MADA.xlsx")
 
+file_exists_check <- file.exists("/Users/andrewruiz/MADA_course/RUIZ-MADA-project/data/raw-data/MOSQ_MADA.xlsx")
+print(file_exists_check)  # Should return TRUE if the path is correct
+print(file_path_mos)
+
+print(file_exists_check)  # This should return TRUE if the file path is correct
 # import the data to R and store it in a dataframe named "mosquito"

### 2024-03-07 products/manuscript/Manuscript.qmd
Fixed manuscript render error
loaded package and inserted image generated in processing qmd

--- 2024-02-23 products/manuscript/Manuscript.qmd
@@
 library(readxl)
+library(dplyr)
+library(kableExtra)
+library(dplyr)
+library(readr)
 
@@
 | Data                         | Description                                                                                                                                           | Source                                |
-|------------------|-------------------------------|-----------------------|
+|------------------------------|-------------------------------------------------------------------------------------------------------------------------------------------------------|---------------------------------------|
 | Mosquito Surveillance Data   | Records of mosquito species counts, locations, and dates of collection.
@@
 
-One factor to consider is if early virus isolations are associated with
-cases in humans and other mammals. The table below shows the date of the
-first EEE virus isolation in mosquitoes and the number of human of other
+It is important to know where the mosquitoes infected with EEE are
+captured. Below is a table showing the virus isolations for EEE in
+trapped mosquitoes by county between 2014 and 2020.
+
+![](images/eee_isolation_county.png){fig-align="center"}
+
+The geographic focus for this study is Bristol County. One factor to
+consider is if early virus isolations are associated with cases in
+humans and other mammals. The table below shows the date of the first
+EEE virus isolation in mosquitoes and the number of human of other
 animal cases by year.
 
-![](images/eee_isolation_cases-01.png){fig-align="center" width="320"}
+![](images/eee_isolation_cases-03.png){fig-align="center"}
 
@@
 ## Basic statistical analysis
+
 *These will be generated in later phases of the exercise*
@@
 
-*These will be generated in later phases of the exercise* 
-# Discussion *this section will be completed when analysis is done*
+*These will be generated in later phases of the exercise* \# Discussion
+*this section will be completed when analysis is done*
 
 ## Strengths and Limitations
+
 *this section will be updated after analysis is complete*
@@
 ## Conclusions *this section will be completed when analysis is done*
+
 {{< pagebreak >}}

### 2024-03-07 products/manuscript/Manuscript.qmd
final edit
removed page break

@@
 | Data                         | Description                                                                                                                                           | Source                                |
-|------------------------------|-------------------------------------------------------------------------------------------------------------------------------------------------------|---------------------------------------|
+|--------------|--------------------------------------------|--------------|
 | Mosquito Surveillance Data   | Records of mosquito species counts, locations, and dates of collection.
@@
 
-## Fullanalysis
+## Full analysis
 
@@
 
-{{< pagebreak >}}
-
 # References

### 2024-03-07 R/processing-code/MADAproject_pt2_ruiz.qmd
commented out absolute path

@@
 
-file_exists_check <- file.exists("/Users/andrewruiz/MADA_course/RUIZ-MADA-project/data/raw-data/MOSQ_MADA.xlsx")
+#file_exists_check <- file.exists("/Users/andrewruiz/MADA_course/RUIZ-MADA-project/data/raw-data/MOSQ_MADA.xlsx")
 print(file_exists_check)  # Should return TRUE if the path is correct

### 2024-03-07 R/eda-code/edacode.R

@@ -4,6 +4,7 @@ library(here) #for data loading/saving
 library(dplyr)
 library(skimr)
 library(ggplot2)
+library(tidyverse)
 
 ## ---- loaddata --------
 #Path to data. Note the use of the here() package and not absolute paths

### 2024-03-08 R/processing-code/MADAproject_pt2_ruiz.qmd
improvements to Part 2
adding more comments and data

@@
 # Introduction
-
 ## This file contains the process for loading, cleaning and exploring the datasets.
-
 # Loading libraries
-
+# this list contains the specific versions used in the script. 
 ```{r, message=FALSE, warning=FALSE}
-library(readxl) #for loading Excel files
-library(httr)   #for getting data from the web
-library(dplyr)  #for data processing/cleaning
-library(tidyr)  #for data processing/cleaning
-library(skimr)  #for visualizing data 
-library(here)   #allows relative pathways for data
-library(readr) #reading and writing data to and from CSV, TSV, and fixed-width files
-library(naniar) #tools to work with missing data
-library(ggplot2) #plot creation
-library(sf) #standardized way to encode spatial data
-library(topicmodels) #tools for topic modeling on text data
-library(openxlsx) #Facilitates reading from and writing to .xlsx files
-library(lubridate) #Simplifies the work with dates and times
-library(knitr) #Allows for dynamic report generation in R
-library(kableExtra) #generates simple HTML or LaTeX tables
-library(webshot2) #tools to take screenshots of web pages 
-library(magick) #advanced image processing in R
-library(leaflet) #Enables the creation of interactive maps
-library(htmlwidgets) #Facilitates the creation of R bindings to JavaScript libraries
+library(readxl) # for loading Excel files (version 1.4.0)
+library(httr) # for getting data from the web (version 1.4.2)
+library(dplyr) # for data processing/cleaning (version 1.0.7)
+library(tidyr) # for data processing/cleaning (version 1.1.3)
+library(skimr) # for visualizing data (version 2.1.3)
+library(here) # allows relative pathways for data (version 1.0.1)
+library(readr) # reading and writing data to and from CSV, TSV, and fixed-width files (version 2.0.1)
+library(naniar) # tools to work with missing data (version 0.6.1)
+library(ggplot2) # plot creation (version 3.3.5)
+library(sf) # standardized way to encode spatial data (version 1.0-2)
+library(topicmodels) # tools for topic modeling on text data (version 0.2-12)
+library(openxlsx) # Facilitates reading from and writing to .xlsx files (version 4.2.4)
+library(lubridate) # Simplifies the work with dates and times (version 1.7.10)
+library(knitr) # Allows for dynamic report generation in R (version 1.33)
+library(kableExtra) # generates simple HTML or LaTeX tables (version 1.3.4)
+library(webshot2) # tools to take screenshots of web pages (version 0.0.2)
+library(magick) # advanced image processing in R (version 2.7.3)
+library(leaflet) # Enables the creation of interactive maps (version 2.0.4.1)
+library(htmlwidgets) # Facilitates the creation of R bindings to JavaScript libraries (version 1.5.4)
 ```
@@
 # Loading data
-
 ## Mosquito trap data
-
-### This dataset contains trap counts by species for all mosquito traps set in the Bristol County from 2007-2023
-
-### this dataset will undergo extensive processing below.
-
+### This dataset contains trap counts by species for all mosquito traps set in the Bristol County from 2007-2023 this dataset will undergo extensive processing below.
 ```{r trap-data-import}
@@
 ## Mosquito trap sites
-
 ### Location for the mosquito trap sites
-
 ```{r}
@@
 trap_location_selected
+
+# Define the path for the trap locations selected dataset RDS
+trap_location_selected_path_rds <- here("data", "processed-data", "rds", "trap_sites_selected.rds")
+
+# Save the trap_location_selected dataframe as an RDS file
+saveRDS(trap_location_selected, trap_location_selected_path_rds)
 ```
@@
 ## Mosquito virus data
-
 ### The virus isolation data is stored in separate spreadsheets by year from 2014 to 2020
-
 ### This code will load all the files, combine them, and then save the combined file in the "raw data" folder
-
 ```{r}
@@
 
+# Define the path for the combined dataset RDS
+combined_path_rds <- here("data", "processed-data", "rds", "combined_virus_data.rds")
+
+# Save the combined data frame as an RDS file
+saveRDS(combined_virus_data, combined_path_rds)
+
 # View the first few rows to confirm it's loaded correctly
@@
 ## Bird data
-
 ### The complete bird dataset is over 600 MB.
@@
 Eight of 50 varibles will be kept and the records will be filtered to show only birds from the Genus Corvus observed in Massachusetts after 2014-01-01
-
 ```{r}
@@
 
+# Define the path for the combined dataset RDS
+bird_path_rds <- here("data", "processed-data", "rds", "filtered_bird.rds")
+
+# Save the bird data frame as an RDS file
+saveRDS(filtered_bird, bird_path_rds)
+
 # ensure the csv loaded properly
@@
 ## Weather data
-
 ### This dataset contains daily weather summaries from the Taunton, MA Airport from 2007-2023
-
 ```{r}
@@
 
+# Define the path for the RDS file
+output_path_wx <- here("data", "processed-data", "rds", "wx_selected.rds")
+
+# Save the wx_selected dataframe as an RDS file
+saveRDS(wx_selected, output_path_wx)
+
 head(wx_selected)
@@
 ## Human and veterinary cases of eastern equine encephalitis and West Nile virus
-
 ```{r}
@@
 cases_human <- read_excel(file_path_human)
+
+# Save 'cases_human' as an RDS file
+output_path_rds_human <- here("data", "processed-data", "rds", "human_cases_processed.rds")
+saveRDS(cases_human, output_path_rds_human)
 
@@
 
+# Save 'cases_vet' as an RDS file
+output_path_rds_vet <- here("data", "processed-data", "rds", "vet_cases_processed.rds")
+saveRDS(cases_vet, output_path_rds_vet)
+
 #load the first few rows of data to verify the data loaded as expected
@@
 # Processing and cleaning the Mosquito trap data
-
 ### Examine the variables and change all column names to lower case
-
 ```{r}
@@
 ### Identify the time period the dataset covers
-
 ```{r}
@@
 We will set the analytical horizon will span 2013 to 2021.*Collection data before 2013 and after 2021 will be removed* along with records with important missing variables identified below.
-
 ### Identify missing data
-
 ```{r}
@@
 This is not a crucial variable, so we can keep those records.The other variable are important and so we will remove those records with the other missing data.
-
 ### The Massachussetts Department of Public Health adpoted a taxonomy change before it was fully vetted.
@@
 To align with accepted standards, I will change them back to *Aedes*.
-
 ```{r}
@@
 ### Execute proposed changes.
-
 ```{r}
@@
 ## Reformat the collection data
-
 ### The Massachusetts Department of Public Health requires that all mosquito control district submit their colelction data in a specific format.
@@
 Even if you sort by species and calculate average count per trap, the average could be inflated since the records will not include trap events with a zero count of that species.
-
 ### In order to correct this, I will pivot the data and create a column for every unique "species code" and sum the "pool size" for each species code.
 The resulting table will have one row for every trap event -a trap event is when "town", "date of collection", "trap type" are all the same.
-
 ```{r}
@@
 print(paste("File saved to:", file_path_wide))
+
+# Define the path for the wide-format dataset RDS
+file_path_wide_rds <- here("data", "processed-data", "rds", "mosquito_wide.rds")
+
+# Save the wide-format data as an RDS file
+saveRDS(wide_data, file_path_wide_rds)
+
 ```
@@
 # Exploring the Data
-
 ##Study location Below is a map of the mosquito trap sites. This is an interactive map and will only work on web or other html-friendly platforms.
-
 ```{r}
@@
 ## EEE in mosquitoes and mammals
-
 ### To give an idea of when EEE virus silled over from mosquitoes and birds and into mammal populations, lets take a look at the data.
@@
 Let's also look at the years that human and veterinary cases occured.
-
 ### From the table, we see that human or othe mammal cases were recorded in 2014, 2018, 2019, and 2020.
@@
 **Because of this, we will pay close attention to 2019 through out the exploration.**
-
 ```{r}
@@
 save_kable(kable_table_isolation_case, file = output_path_png)
+
+# Define the paths for the RDS files
+first_positive_eee_test_by_year_path <- here("data", "processed-data", "rds", "first_positive_eee_test_by_year.rds")
+human_cases_by_year_path <- here("data", "processed-data", "rds", "human_cases_by_year.rds")
+animal_cases_by_year_path <- here("data", "processed-data", "rds", "animal_cases_by_year.rds")
+combined_eee_data_path <- here("data", "processed-data", "rds", "combined_eee_data.rds")
+
+# Save the data frames as RDS files
+saveRDS(first_positive_eee_test_by_year, first_positive_eee_test_by_year_path)
+saveRDS(human_cases_by_year, human_cases_by_year_path)
+saveRDS(animal_cases_by_year, animal_cases_by_year_path)
+saveRDS(combined_eee_data, combined_eee_data_path)
 ```
@@
 ## Which county has the most EEE positive mosquitoes of those tested?
-
 ```{r}
@@
 #output_path_qmd <- here("results", "tables", "eee_isolation_county.html")
-output_path_county <- here("results", "tables", "eee_isolation_county.png")
+output_path_county <- here("results", "tables", "eee_isolation_county.html")
 
@@
 Using the new mosquito collection table created from the pivot, let's calculate the proportion of MEL captured compared to all other mosquito vector species and create a plot to visualize it.
+```{r}
+library(dplyr)
+library(ggplot2)
+library(readr) # For read_csv
+library(here)
 
-###
-
-```{r}
 # Define path to the data file
-trap_wide_path = here("data", "processed-data", "mosquito_wide.csv")
+trap_wide_path <- here("data", "processed-data", "mosquito_wide.csv")
 
 # Read the data
-trap_proportion = read.csv(trap_wide_path)
+trap_proportion <- read_csv(trap_wide_path) %>%
+  mutate(Year = year(as.Date(`collection date`, format = "%Y-%m-%d")))
 
-# Convert 'collection.date' to Date type using the correct format and extract the year
-trap_proportion$Year <- year(ymd(trap_proportion$collection.date))
+# Calculate the proportion of MEL out of the total for all species
+proportion_mel_by_year <- trap_proportion %>%
+  group_by(Year) %>%
+  summarise(Total_MEL = sum(MEL, na.rm = TRUE),
+            Total_All = sum(MEL + PER + PIP + RES + SAL + TRI + SAP, na.rm = TRUE),
+            MEL_Prop = Total_MEL / Total_All)
 
-# Sum all specified species by the calculated year, including any additional species codes you have
-yearly_sums <- trap_proportion %>%
+# Calculate the average number of MEL by year
+average_mel_by_year <- trap_proportion %>%
   group_by(Year) %>%
-  summarise(across(c("MEL", "PER", "PIP", "RES", "SAL", "TRI", "SAP"), sum, na.rm = TRUE)) %>%
-  # After summarising, calculate the total counts for all species per year
-  mutate(Total = rowSums(across(c("MEL", "PER", "PIP", "RES", "SAL", "TRI", "SAP"))),
-         # Calculate the proportion of MEL out of the total for all species
-         MEL_Prop = MEL / Total)
+  summarise(Average_MEL = mean(MEL, na.rm = TRUE))
 
-# View the result with the proportion of MEL
-print(yearly_sums)
+# Save the processed data as RDS and CSV
+saveRDS(proportion_mel_by_year, here("data", "processed-data", "rds", "yearly_mel_proportion.rds"))
+write.csv(proportion_mel_by_year, here("data", "processed-data", "yearly_mel_proportion.csv"), row.names = FALSE)
 
-library(ggplot2)
+saveRDS(average_mel_by_year, here("data", "processed-data", "rds", "yearly_mel_average.rds"))
+write.csv(average_mel_by_year, here("data", "processed-data", "yearly_mel_average.csv"), row.names = FALSE)
 
-mel_plot = ggplot(yearly_sums, aes(x = factor(Year), y = MEL_Prop, fill = "MEL")) +
-  geom_bar(stat = "identity", color = "black") +
-  labs(title = expression(Proportion ~ of ~ italic("Culiseta melanura") ~ (MEL) ~ Mosquitoes ~ by ~ Year),
-       x = NULL,
+# Plot for the proportion of MEL by year
+ggplot(proportion_mel_by_year, aes(x = factor(Year), y = MEL_Prop)) +
+  geom_col(fill = "steelblue") +
+  labs(title = "Proportion of Culiseta melanura (MEL) by Year",
+       x = "Year",
        y = "Proportion of MEL") +
-  scale_fill_discrete(name = "Species") +
   theme_minimal() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-mel_plot
-# Specify the file path where you want to save the plot
-#file_path_mel = here("results", "figures", "mel_proportion.png")
-# Use ggsave to save the plot
-#ggsave(file_path_mel, plot = mel_plot, width = 8, height = 6, dpi = 300)
+
+ggsave(here("results", "figures", "mel_proportion_by_year.png"), width = 8, height = 6, dpi = 300)
+
+# Plot for the average count of MEL by year
+mel_plot = ggplot(average_mel_by_year, aes(x = factor(Year), y = Average_MEL, fill = "Year")) +
+  geom_col() +
+  labs(title = "Average Count of Culiseta melanura (MEL) by Year",
+       x = "Year",
+       y = "Average Count of MEL") +
+  scale_fill_viridis_d() +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+ggsave(here("results", "figures", "average_mel_by_year.png"), plot = mel_plot, width = 8, height = 6, dpi = 300)
 
@@
 ```{r annual precip}
-
 # Convert DATE from character to Date type using as.Date and specifying the format
@@
 
+# Save the filtered weather data as an RDS file
+filtered_wx_path_rds <- here("data", "processed-data", "rds", "filtered_wx_2013_2020.rds")
+saveRDS(filtered_wx, filtered_wx_path_rds)
+
 # Sum PRCP by Year for the filtered data
@@
   summarise(Total_PRCP = sum(PRCP, na.rm = TRUE))
+
+# Save the summarized precipitation data as an RDS file
+summed_prcp_by_year_path_rds <- here("data", "processed-data", "rds", "summed_prcp_by_year_filtered_2013_2020.rds")
+saveRDS(summed_prcp_by_year_filtered, summed_prcp_by_year_path_rds)
 
@@
 ```
+### Now I want to plot the ave mel by year with the total precip by year
+
+```{r}
+
+# Load the datasets
+yearly_mel_ave <- readRDS(here("data", "processed-data", "rds", "yearly_mel_ave.rds"))
+summed_prcp_by_year_filtered <- readRDS(here("data", "processed-data", "rds", "summed_prcp_by_year_filtered_2013_2020.rds"))
+
+# Combine the datasets
+combined_data <- left_join(yearly_mel_ave, summed_prcp_by_year_filtered, by = "Year")
+
+# Define the path for the RDS file with a descriptive name
+combined_data_path_rds <- here("data", "processed-data", "rds", "combined_mel_avg_and_total_precip_2013_2020.rds")
+
+# Save the combined data as an RDS file
+saveRDS(combined_data, combined_data_path_rds)
+
+ggplot(data = combined_data) +
+  geom_col(aes(x = factor(Year), y = Average_MEL), fill = "blue") +
+  geom_line(aes(x = factor(Year), y = Total_PRCP, group = 1), color = "red", linetype = "dashed") +
+  geom_point(aes(x = factor(Year), y = Total_PRCP), color = "red", size = 3) +
+  geom_hline(yintercept = 45.6, linetype = "dotted", color = "darkgreen", size = 1) +
+  annotate("text", x = Inf, y = 45.6, label = "Avg Precip: 45.6 inches", 
+           hjust = 3.2, vjust = -1.1, color = "darkgreen", size = 3.5, fontface = "bold", angle = 0) +
+  scale_y_continuous(name = "Average Count of MEL", 
+                     sec.axis = sec_axis(~ ., name = "Total Precipitation (inches)")) +
+  labs(title = "Average Count of MEL Mosquitoes and Total Precipitation by Year",
+       x = "Year") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
+        axis.title.x = element_text(face = "bold"))
+# Save the plot
+ggsave(filename = here("results", "figures", "combined_trends_mel_precip.png"), plot = last_plot(), width = 10, height = 6, dpi = 300)
+
+```

### 2024-03-08 data/raw-data/bristol_cty.qmd
improvements to Part 2
adding more comments and data

<!DOCTYPE qgis PUBLIC 'http://mrcc.com/qgis.dtd' 'SYSTEM'>
<qgis version="3.36.0-Maidenhead">
  <identifier></identifier>
  <parentidentifier></parentidentifier>
  <language></language>
  <type>dataset</type>
  <title></title>
  <abstract></abstract>
  <links/>
  <dates/>
  <fees></fees>
  <encoding></encoding>
  <crs>
    <spatialrefsys nativeFormat="Wkt">
      <wkt>GEOGCRS["NAD83",DATUM["North American Datum 1983",ELLIPSOID["GRS 1980",6378137,298.257222101,LENGTHUNIT["metre",1]]],PRIMEM["Greenwich",0,ANGLEUNIT["degree",0.0174532925199433]],CS[ellipsoidal,2],AXIS["geodetic latitude (Lat)",north,ORDER[1],ANGLEUNIT["degree",0.0174532925199433]],AXIS["geodetic longitude (Lon)",east,ORDER[2],ANGLEUNIT["degree",0.0174532925199433]],USAGE[SCOPE["Geodesy."],AREA["North America - onshore and offshore: Canada - Alberta; British Columbia; Manitoba; New Brunswick; Newfoundland and Labrador; Northwest Territories; Nova Scotia; Nunavut; Ontario; Prince Edward Island; Quebec; Saskatchewan; Yukon. Puerto Rico. United States (USA) - Alabama; Alaska; Arizona; Arkansas; California; Colorado; Connecticut; Delaware; Florida; Georgia; Hawaii; Idaho; Illinois; Indiana; Iowa; Kansas; Kentucky; Louisiana; Maine; Maryland; Massachusetts; Michigan; Minnesota; Mississippi; Missouri; Montana; Nebraska; Nevada; New Hampshire; New Jersey; New Mexico; New York; North Carolina; North Dakota; Ohio; Oklahoma; Oregon; Pennsylvania; Rhode Island; South Carolina; South Dakota; Tennessee; Texas; Utah; Vermont; Virginia; Washington; West Virginia; Wisconsin; Wyoming. US Virgin Islands. British Virgin Islands."],BBOX[14.92,167.65,86.46,-47.74]],ID["EPSG",4269]]</wkt>
      <proj4>+proj=longlat +datum=NAD83 +no_defs</proj4>
      <srsid>3401</srsid>
      <srid>4269</srid>
      <authid>EPSG:4269</authid>
      <description>NAD83</description>
      <projectionacronym>longlat</projectionacronym>
      <ellipsoidacronym>EPSG:7019</ellipsoidacronym>
      <geographicflag>true</geographicflag>
    </spatialrefsys>
  </crs>
  <extent/>
</qgis>

### 2024-03-12 R/processing-code/MADAproject_pt2_ruiz.qmd
Updates to pt 2.
created rds files for use in later updates.

@@
 # Introduction
+
 ## This file contains the process for loading, cleaning and exploring the datasets.
 # Loading libraries
-# this list contains the specific versions used in the script. 
+# This list contains the specific versions used in the script.
 ```{r, message=FALSE, warning=FALSE}
@@
 library(htmlwidgets) # Facilitates the creation of R bindings to JavaScript libraries (version 1.5.4)
+library(suncalc) # Calculates sunset and sunrise times based on location (version 0.5.1)
 ```
@@
 ### This dataset contains trap counts by species for all mosquito traps set in the Bristol County from 2007-2023 this dataset will undergo extensive processing below.
+#### The dataset was obtained though and open records request with the Commonwealth of MA.
 ```{r trap-data-import}
@@
 ## Mosquito trap sites
+
 ### Location for the mosquito trap sites
+
 ```{r}
@@
 ## Mosquito virus data
+
 ### The virus isolation data is stored in separate spreadsheets by year from 2014 to 2020
+
 ### This code will load all the files, combine them, and then save the combined file in the "raw data" folder
+
 ```{r}
@@
 ## Bird data
+
 ### The complete bird dataset is over 600 MB.
@@
 ## Human and veterinary cases of eastern equine encephalitis and West Nile virus
+
 ```{r}
@@
 We will set the analytical horizon will span 2013 to 2021.*Collection data before 2013 and after 2021 will be removed* along with records with important missing variables identified below.
+
 ### Identify missing data
+
 ```{r}
@@
 This is not a crucial variable, so we can keep those records.The other variable are important and so we will remove those records with the other missing data.
+
 ### The Massachussetts Department of Public Health adpoted a taxonomy change before it was fully vetted.
@@
 Even if you sort by species and calculate average count per trap, the average could be inflated since the records will not include trap events with a zero count of that species.
+
 ### In order to correct this, I will pivot the data and create a column for every unique "species code" and sum the "pool size" for each species code.
 The resulting table will have one row for every trap event -a trap event is when "town", "date of collection", "trap type" are all the same.
+
 ```{r}
@@
 # Exploring the Data
+
 ##Study location Below is a map of the mosquito trap sites. This is an interactive map and will only work on web or other html-friendly platforms.
+
 ```{r}
@@
 **Because of this, we will pay close attention to 2019 through out the exploration.**
+
 ```{r}
@@
 
-# Define the path where you want to save the HTML file
+# Define the path where you want 
 #output_path_quarto <- here("results", "tables", "eee_isolation_cases")
@@
 
-# Save the kableExtra table as an HTML file
 #save_kable(kable_table_isolation_case, file = output_path_quarto)
@@
 ## Which county has the most EEE positive mosquitoes of those tested?
+
 ```{r}
@@
 Using the new mosquito collection table created from the pivot, let's calculate the proportion of MEL captured compared to all other mosquito vector species and create a plot to visualize it.
+
 ```{r}
@@
 ```
+
 ### Now I want to plot the ave mel by year with the total precip by year
-
 ```{r}
@@
 ggsave(filename = here("results", "figures", "combined_trends_mel_precip.png"), plot = last_plot(), width = 10, height = 6, dpi = 300)
+```
 
+```{r}
+# Ensure the suncalc library is loaded
+library(suncalc)
+get_sun_times <- function(date, latitude, longitude) {
+  times <- getSunlightTimes(date = date, lat = latitude, lon = longitude, tz = "UTC")
+
+  # Convert sunrise and sunset times to POSIXct directly without checking class
+  # Assume times are returned in a standard format that can be converted directly
+  sunrise <- as.POSIXct(times$sunrise, format = "%Y-%m-%dT%H:%M:%S", tz = "UTC")
+  sunset <- as.POSIXct(times$sunset, format = "%Y-%m-%dT%H:%M:%S", tz = "UTC")
+
+  # Format sunrise and sunset times
+  formatted_sunrise <- format(sunrise, "%H:%M:%S", tz="UTC")
+  formatted_sunset <- format(sunset, "%H:%M:%S", tz="UTC")
+
+  return(c(as.character(date), formatted_sunrise, formatted_sunset))
+}
+
+# Define the geographical coordinates for Bristol County, MA (latitude and longitude)
+latitude <- 41.8744
+longitude <- -71.0166
+
+# Generate a sequence of dates from 2015-01-01 to 2024-12-31
+dates <- seq(as.Date("2015-01-01"), as.Date("2024-12-31"), by = "day")
+
+# Calculate sunrise and sunset times for each date
+sun_times <- lapply(dates, function(date) {
+  get_sun_times(date, latitude, longitude)
+})
+
+# Convert the list to a data frame
+sun_times_df <- do.call(rbind, sun_times)
+colnames(sun_times_df) <- c("Date", "Sunrise", "Sunset")
+
+# Write the data frame to a CSV file
+write.csv(sun_times_df, file = "sunrise_sunset_times_bristol_county_ma.csv", row.names = FALSE)
+
+# Define the file path where you want to save the sunrise and sunset times data frame
+sunrise_sunset_times_path <- here("data", "processed-data", "rds", "sunrise_sunset_times_bristol_county_ma.rds")
+
+# Save the sun_times_df data frame as an RDS file
+saveRDS(sun_times_df, sunrise_sunset_times_path)
+
+# Output to confirm file saving
+cat("Sunrise and sunset times saved to:", sunrise_sunset_times_path, "\n")
+
+#moon
+# Generate a sequence of dates from 2015-01-01 to 2022-12-31
+dates <- seq(as.Date("2015-01-01"), as.Date("2022-12-31"), by = "day")
+
+# Function to get moon phase for a given date
+get_moon_phase <- function(date) {
+  illumination <- getMoonIllumination(date)
+  phase <- illumination$phase
+  return(phase)
+}
+
+# Calculate moon phase for each date
+moon_phases <- sapply(dates, get_moon_phase)
+
+# Create a data frame with dates and their corresponding moon phases
+moon_phases_df <- data.frame(Date = dates, MoonPhase = moon_phases)
+
+# View the first few rows of the data frame
+head(moon_phases_df)
 ```

### 2024-03-12 products/manuscript/Manuscript.qmd
manuscript update
saving manuscript edits. began restructuring the sections.

@@
 
-This project will lay the foundation for my dissertation research. **The
-goal for this phase of the project is to answer which environmental,
-ecological, and temporal factors influence the transmission cycles for
-EEE in southeastern Massachusetts.**
+# Problem statement
+
+Which environmental,ecological, and temporal factors influence the
+transmission cycles for EEE in southeastern Massachusetts and which
+models best describe the relationship between these factors.
+
+This project will lay the foundation for dissertation research that will
+begin in June, 2024. The research will focus on three
 
@@
 
-## General Background Information
+## Background
 
@@
 population through pesticide use and environmental modifications like
-removing standing water. The Bristol County Mosquito Control Project
-(BCMCP) coordinates mosquito surveillance and testing in the county from
-June to October. BCMCP has used the same sentinel collection sites and
-trapping methods for over 20 years. Once trapped, mosquitoes are sorted
-by species, and vector species are submitted to the Massachusetts
-Department of Health’s (MDPH) State Lab for PCR virus testing. Results
-are available within 24 hours of submission. When rates of EEE mosquito
-infections are above a defined threshold, MDPH notifies the local boards
-of health and recommends preventive measures. Recommended measures
-include outreach and education to increase personal protective practices
-. Additionally, MDPH may recommend that cities and towns in the affected
-areas cancel evening outdoor events and discourage outdoor activities
-when mosquito vectors are most active (dawn and dusk). When infection
-rates are high enough to indicate imminent human transmission, the
-Commonwealth of Massachusetts will recommend and fund aerial pesticide
-applications over the affected areas.
+removing standing water.
 
-## Description of data and data source
+The Bristol County Mosquito Control Project (BCMCP) coordinates mosquito
+surveillance and testing in the county from June to October. BCMCP has
+used the same sentinel collection sites and trapping methods for over 20
+years. Once trapped, mosquitoes are sorted by species, and vector
+species are submitted to the Massachusetts Department of Health’s (MDPH)
+State Lab for PCR virus testing. Results are available within 24 hours
+of submission. When rates of EEE mosquito infections are above a defined
+threshold, MDPH notifies the local boards of health and recommends
+preventive measures. Recommended measures include outreach and education
+to increase personal protective practices . Additionally, MDPH may
+recommend that cities and towns in the affected areas cancel evening
+outdoor events and discourage outdoor activities when mosquito vectors
+are most active (dawn and dusk). When infection rates are high enough to
+indicate imminent human transmission, the Commonwealth of Massachusetts
+will recommend and fund aerial pesticide applications over the affected
+areas.
 
-*The cornerstone of this research will be the mosquito collection data
-provided by a mosquito control project in SE Massachusetts. A formal
-data-sharing agreement will be in place by January 2024. The dataset
-encompasses nearly two decades of detailed mosquito surveillance. The
-datasets include spatial identifiers. Mosquito collection data will be
-integrated with other datasets using spatial or temporal attributes.
-These datasets will include biological and ecological data,
-environmental and spatial data, and demographic data, which are crucial
-in understanding the transmission dynamics of EEE. Potential datasets
-with their sources are outlined in the table below. Other datasets may
-be included based on availability and need.*
+# Methods
+
+## Study area
+
+Bristol County, Massachusetts covers an area of 691 square miles with a
+population density of 962 people per square mile. It borders Rhode
+Island to the west, Plymouth County, MA to the east, and Norfolk County,
+MA to the north.The county is situated on Buzzard's Bay to the south.
+
+Atlantic White Cedar (AWC) swamps span over 4,300 acres within
+Massachusetts, primarily concentrated in Plymouth and Bristol counties.
+These swamps serve as critical breeding habitats for Culiseta melanura
+(MEL), which is recognized as the primary enzootic vector for Eastern
+Equine Encephalitis (EEE). AWC trees have large roots that form "crypts"
+that hold water throughout the year. These crypts serve as a main
+habitat for MEL larval growth.
+
+Between 2003 and 2022 Plymouth and Bristol counties were the epicenters
+for EEE transmission in Massachusetts, accounting for 24 out of the
+state's 41 human EEE cases.The Bristol County Mosquito Control Project
+(BCMCP) conducts mosquito surveillance and abatement activities for the
+county and often coordinates activities with Plymouth County Mosquito
+Control Project.
+
+## Data collection and processing
+
+The cornerstone of this project is mosquito surveillance data. This
+dataset includes mosquito counts by species for all mosquito traps set
+in Bristol County, MA between 2007 to 2023. In addition to mosquito
+counts by species, the dataset includes the town or city in which they
+were collected, the collection data, and whether the mosquitoes were
+submitted for virus testing.
+
+While the surveillance data provide insight into mosquito population
+estimates and changes in species composition, it is missing one key
+feature. The virus isolation results are not included in this dataset.
+However, virus isolations in mosquitoes are published on the MDPH
+website by town/city and collection date. With this information, one can
+identify the trap event contained in the dataset which allow for
+calculations of key measure, such as minimum infection rate (MIR).
+
+The surveillance dataset is structured to conform with state and federal
+reporting requirements. As a result, the format of the data must be
+transformed to in order to make it ready for analysis. Once the data
+structure is corrected, it can be joined to other relevant datasets
+using spatial or temporal attributes.
+
+This dataset will be integrated with other datasets using spatial or
+temporal attributes. These datasets will include biological and
+ecological data, environmental and spatial data, and demographic data,
+which are crucial in understanding the transmission dynamics of EEE.
+Potential datasets with their sources are outlined in the table below.
+Other datasets may be included based on availability and need.
 
 | Data                         | Description                                                                                                                                           | Source                                |
-|--------------|--------------------------------------------|--------------|
+|-------------------|----------------------------------|-------------------|
 | Mosquito Surveillance Data   | Records of mosquito species counts, locations, and dates of collection.
@@
 environmental, ecological, and temporal factors influence the
-transmission cycles for EEE in southeastern Massachusetts.** While the
-following questions will not be addressed in this phase, this project
-will hopefully inform research that will answer the following questions:
+transmission cycles for EEE in southeastern Massachusetts.**
+
+While the following questions will not be addressed in this phase, this
+project will hopefully inform research that will answer the following
+questions:
 

### 2024-03-14 R/processing-code/MADAproject_pt2_ruiz.qmd
Pt 3
updated qmd and manuscript

@@
 library(suncalc) # Calculates sunset and sunrise times based on location (version 0.5.1)
+library(stringr) # Makes string manipulation easier and more consistent (version 1.5.1)
+library(ISOweek) # Calculates week number according to ISO 8601 standards (version 0.6-2)
 ```
@@
 head(trap_location)
-str(trap_location)
 
@@
 # there is not a need to delete the records
-#However, we will delete the columns
-
+# this is a free text field that provides notes about the location.
+# they are not needed so we will delete them
 trap_location_selected <- trap_location %>%
@@
 ## Mosquito virus data
-
+### The data includes all the positive PCR results for EEE and WNV for all mosquitoes submitted for testing
+### These files were obtained from: https://www.mass.gov/lists/arbovirus-surveillance-plan-and-historical-data
 ### The virus isolation data is stored in separate spreadsheets by year from 2014 to 2020
-
 ### This code will load all the files, combine them, and then save the combined file in the "raw data" folder
-
 ```{r}
@@
 ## Weather data
-
+### temperature and precipitation are important factors in mosquito population dynamics
 ### This dataset contains daily weather summaries from the Taunton, MA Airport from 2007-2023
-
 ```{r}
@@
 ## Human and veterinary cases of eastern equine encephalitis and West Nile virus
-
+### these were extracted from annual mosquito reports publised here: https://www.mass.gov/lists/arbovirus-surveillance-plan-and-historical-data
+### human cases are listed by county only. animal cases are listed by town/city (with the exception of 1 year)
 ```{r}
@@
 However, the virus and case data only covers the years 2014 to 2020.
-Because mosquito populations can be influences by processes of the previous year.
+Because mosquito populations can be influenced by processes of the previous year.
 We will set the analytical horizon will span 2013 to 2021.*Collection data before 2013 and after 2021 will be removed* along with records with important missing variables identified below.
@@
 ### Identify missing data
-
 ```{r}
@@
 unique(mosquito_raw$genus)
-
 ```
@@
 
-# Change records where genus is "Ochlerotatus" to "Aedes"
-#mosquito_modified$genus[mosquito_modified$genus == "Ochlerotatus"] <- "Aedes"
+# Change records where genus is "Ochlerotatus" to "Aedes" to adhere to current taxonomy standards
+mosquito_modified$genus[mosquito_modified$genus == "Ochlerotatus"] <- "Aedes"
 
-# Filter out records where Collection Date is before 2013 or after 2021
-# Ensure Collection Date is in Date format if it's not already
+# Ensure Collection Date is in Date format 
 mosquito_modified$`collection date` <- as.Date(mosquito_modified$`collection date`)
@@
 # Convert all species code values to uppercase
+# This will minimize issues were data as entered incorrectly
+# for example Mel, mel, and MEL
 mosquito_modified <- mosquito_modified %>%
@@
 
-# Now apply the date filter
-mosquito_modified <- mosquito_modified[mosquito_modified$`collection date` >= as.Date("2013-01-01") & 
-                                       mosquito_modified$`collection date` <= as.Date("2021-12-31"), ]
-
 # Remove rows with NA values in the remaining columns of the modified dataframe, 
 # except for "Submitted for Testing" which we'll handle separately
+# when Submitted for testing is NA or missing, it indicates that there were no mosquitoes captured during a trap event.
+# This is common at the beginning and end of season and also when a trap malfunctions
 temp_columns <- names(mosquito_modified)
 temp_columns <- temp_columns[temp_columns != "submitted for testing"]
+mosquito_modified <- na.omit(mosquito_modified, cols = temp_columns)
 
-# Applying NA omit on selected columns only
-mosquito_clean_except_submitted <- na.omit(mosquito_modified[temp_columns])
+# Add "week_num" column
+# This code ensures that the ISO 8601 standard is applied to the week number calculation
+# Week 1 of any year is the week that contains January 4th, or equivalently, it's the week that contains the first Thursday of January.
+# Weeks start on Monday and end on Sunday.
+# The last week of the year, week 52 or 53, is the one that contains December 28th.
+# we will use week_num to join tables
+# the format will be YYYY-WW. Week 8 of 2019 will look like: 2019-08
+mosquito_modified <- mosquito_modified %>%
+  mutate(
+    # Directly format the collection date into "YYYY-WW"
+    week_num = paste0(year(`collection date`), "-", 
+                      sprintf("%02d", isoweek(`collection date`)))
+  )
 
-# Since na.omit on selected columns messes with row alignment, let's filter the original dataset based on the cleaned IDs
-clean_ids <- row.names(mosquito_clean_except_submitted)
-mosquito_final <- mosquito_modified[clean_ids, ]
+# before filtering the dates to have a cleaned dataset with all the records.
+write_csv(mosquito_modified, here("data", "processed-data", "mosquito_clean_all.csv"))
+saveRDS(mosquito_modified, here("data", "processed-data", "rds", "mosquito_clean_all.rds"))
 
-# Add back "Submitted for Testing" from the original dataframe if needed
-# Note: This step might be redundant if "Submitted for Testing" was not actually removed or if it's included in the cleaning process
-mosquito_final$`submitted for testing` <- mosquito_raw$`submitted for testing`[clean_ids]
+# Now apply the date filter to match the analytical horizon of the project
+mosquito_final <- mosquito_modified %>%
+  filter(`collection date` >= as.Date("2013-01-01") & `collection date` <= as.Date("2021-12-31"))
 
@@
 ## Reformat the collection data
-
-### The Massachusetts Department of Public Health requires that all mosquito control district submit their colelction data in a specific format.
+### The Massachusetts Department of Public Health and US CDC require that all mosquito control districts submit their colelction data in a specific format.
 This format divides each trap event into separate rows based on mosquito species.
@@
 Even if you sort by species and calculate average count per trap, the average could be inflated since the records will not include trap events with a zero count of that species.
-
 ### In order to correct this, I will pivot the data and create a column for every unique "species code" and sum the "pool size" for each species code.
 The resulting table will have one row for every trap event -a trap event is when "town", "date of collection", "trap type" are all the same.
-
+### The resulting data frame will rectangle in shape and contain 1 row per trap event with all possible species as a separate column. 
 ```{r}
-library(dplyr)
-library(tidyr)
-
 # Group and summarize the mosquito data to prepare for wide format transformation.
@@
 ## EEE in mosquitoes and mammals
-
-### To give an idea of when EEE virus silled over from mosquitoes and birds and into mammal populations, lets take a look at the data.
+### To give an idea of when EEE virus spilled over from mosquitoes and birds and into mammal populations, lets take a look at the data.
 Let's see when the first EEE isolations were found in mosquitoes each year.
 Let's also look at the years that human and veterinary cases occured.
-
 ### From the table, we see that human or othe mammal cases were recorded in 2014, 2018, 2019, and 2020.
@@
 Using the new mosquito collection table created from the pivot, let's calculate the proportion of MEL captured compared to all other mosquito vector species and create a plot to visualize it.
-
 ```{r}
@@
 ## Mosquito abundance is affected by weather. Mosquitoes depend on standing water to complete their larval development. Let's take a look at the annual precipitation using the wx data frame.
-
 ```{r annual precip}

### 2024-03-14 products/manuscript/Manuscript.qmd
Pt 3
updated qmd and manuscript

@@
 Massachusetts, primarily concentrated in Plymouth and Bristol counties.
-These swamps serve as critical breeding habitats for Culiseta melanura
+These swamps serve as critical breeding habitats for *Culiseta melanura*
 (MEL), which is recognized as the primary enzootic vector for Eastern
@@
 
-The cornerstone of this project is mosquito surveillance data. This
-dataset includes mosquito counts by species for all mosquito traps set
-in Bristol County, MA between 2007 to 2023. In addition to mosquito
+The mosquito surveillance dataset encompasses counts by species from all mosquito traps deployed in Bristol County, MA, from 2007 to 2023.
+The BCMCP, perform mosquito surveillance from June through October annually.
+During this interval, CDC light traps baited with carbon dioxide are placed at various sentinel sites each week.
+Some of the sentinel site have been used for over 50 years.
+The traps are installed in the afternoon and retrieved the following morning.
+Designed to attract mosquitoes in search of blood meals, light traps serve as a reliable measure of the actively feeding mosquito population, thereby establishing them as the standard method for EEE surveillance.
+
+Additionally, the county employs gravid traps, which are specifically aimed at attracting and capturing gravid *Culex* mosquitoes looking for egg-laying sites.
+While these traps are primarily used for West Nile Virus (WNV) surveillance, they hold significant value for EEE monitoring as well.
+Given the role of *Culex* species as bridge vectors for EEE, their capture in gravid traps provides critical insight into the risk of virus transmission.
+
+The  In addition to mosquito
 counts by species, the dataset includes the town or city in which they
@@
 environmental, ecological, and temporal factors influence the
-transmission cycles for EEE in southeastern Massachusetts.**
+transmission cycles for EEE in southeastern Massachusetts.** From prior
+research we know that the mosquito life cyle is influenced by a number
+of factors. Temperature plays a major role in the mosquito life cyle
+virus transmission dynamics. As temperature increases, the rate of
+larval development also increases and the time between when a mosquito
+ingests virus in a bloodmeal and when it can transmit it decreases. The
+result is that the mosquito population can grow at faster rate and they
+will have more time to transmit the virus. Precipitation increases the
+amount of standing water and increases the chances of larval survival
+during the mosquito season and also during the winter months for those
+species that overwinter in the larval stage. For EEE, the main enzootic
+vector, MEL, overwinters as larvae.
+
+While these relationships are well known, they are not well described
+for EEE, especially in terms of quantifying the role these two factors
+play in EEE transmission. Specifically, there's a gap in our
+understanding of how these factors influence the number of mosquitoes
+caught in traps and the rate of virus infection among those mosquitoes.
+This gap is more pronounced when distinguishing the impacts on enzootic
+(mosquitoes that primarily feed on birds) versus bridge vector
+(mosquitoes that can transmit the virus from birds to humans) mosquito
+species. Even less well described is the effect the factors have on
+enzootic vs. bridge vector mosquito species.
+
+Moreover, the influence of other variables, such as bird populations,
+remains even less explored. Birds, particularly species of passerines,
+serve as amplifying hosts for EEEV and WNV, potentially affecting the
+level of virus present in an area. Despite their importance in the virus
+transmission cycle, the effect of fluctuations in bird populations on
+the prevalence of these viruses throughout a season has received minimal
+attention. Finally, given the migratory nature of many passerine
+species, it's worth exploring whether EEEV activity in the warmer
+regions of the United States, during the months preceding mosquito
+season in Massachusetts, could serve as an early indicator of virus
+activity in the state. Understanding these dynamics is crucial for
+predicting outbreaks and implementing effective control measures.
 

### 2024-03-15 R/processing-code/MADAproject_pt2_ruiz.qmd
Part 3
Continued work on EDA. Most work for part 3 is contained in the R/eda-code folder. R code will eventually be pulled into the qmd via rds files. Continuing to update the manuscript and add better descriptions to code to facilitate better replication. Still work in progress. All read me files have been updated and will need further attention as the project moves forward.

@@
 # Load complete bird data
+# this uses an absolute path because the file is too large to store on github. 
+# the filtered_bird_data.csv file will be the one used for any analysis
 bird <- read.table("/Users/andrewruiz/0011508-240216155721649.csv", sep = "\t", header = TRUE, fill = TRUE)
@@
 
+wx <- wx %>%
+  mutate(
+    DATE = mdy(DATE),  # Parse the DATE column from "month/day/year"
+    week_num = paste0(year(DATE), "-", sprintf("%02d", isoweek(DATE)))  # Create week_num
+  )
 #load the first few rows of data to verify the data loaded as expected
@@
 wx_selected <- wx %>%
-  select(STATION, NAME, DATE, PRCP)
+  select(STATION, NAME, DATE, PRCP, TMAX, week_num)
 
@@
 output_path_wx <- here("data", "processed-data", "rds", "wx_selected.rds")
-
+output_csv_wx <- here("data", "processed-data", "wx_selected.csv")
 # Save the wx_selected dataframe as an RDS file
 saveRDS(wx_selected, output_path_wx)
-
+write.csv(wx_selected, output_csv_wx, row.names = FALSE)
 head(wx_selected)
@@
 #### The mosquito dataset spans 2007 to 2023.
-However, the virus and case data only covers the years 2014 to 2020.
-Because mosquito populations can be influenced by processes of the previous year.
-We will set the analytical horizon will span 2013 to 2021.*Collection data before 2013 and after 2021 will be removed* along with records with important missing variables identified below.
-
+However, the virus and case data only covers the years 2012 to 2020.
+We will create two datasets One that filters the dates to cover the virus and case data.
+And one with all records.
 ### Identify missing data
@@
   mutate(
-    # Directly format the collection date into "YYYY-WW"
-    week_num = paste0(year(`collection date`), "-", 
-                      sprintf("%02d", isoweek(`collection date`)))
+    week_num = paste0(year(`collection date`), "-", sprintf("%02d", isoweek(`collection date`)))
   )
+# Check the structure of the final cleaned dataframe
+str(mosquito_modified)
 
@@
 saveRDS(mosquito_modified, here("data", "processed-data", "rds", "mosquito_clean_all.rds"))
-
-# Now apply the date filter to match the analytical horizon of the project
-mosquito_final <- mosquito_modified %>%
-  filter(`collection date` >= as.Date("2013-01-01") & `collection date` <= as.Date("2021-12-31"))
-
-# Check the structure of the final cleaned dataframe
-str(mosquito_final)
 ```
@@
 # while excluding 'submitted for testing' from the grouping to avoid potential inconsistencies.
-mosquito_summarized <- mosquito_final %>%
+mosquito_summarized <- mosquito_modified %>%
   group_by(`collection date`, town, `trap type`, `species code`, mcd, `submitted for testing`) %>%
@@
    # Ensure that the summarization does not drop any groups by setting .groups = 'drop'.
+mosquito_summarized <- mosquito_modified %>%
+  group_by(`collection date`, week_num, town, `trap type`, `species code`, mcd, `submitted for testing`) %>%
+  summarise(pool_size_sum = sum(`pool size`, na.rm = TRUE), .groups = 'drop')
 
@@
     values_from = pool_size_sum,
-    values_fill = list(pool_size_sum = 0),
-    id_cols = c(`collection date`, town, `trap type`, mcd)
-  )
+    id_cols = c(`collection date`, week_num, town, `trap type`, mcd)
+  ) %>%
+  # Now use replace_na() to fill in NAs, specifying columns dynamically or by name after checking what columns are created
+  replace_na(list(your_dynamic_or_specific_column_names_here = 0))
 
@@
 # Specify the file path using here()
-file_path_wide <- here("data", "processed-data", "mosquito_wide.csv")
+file_path_wide <- here("data", "processed-data", "mosquito_wide_all.csv")
 
@@
 
-# Print the file path to confirm where the file has been saved
-print(paste("File saved to:", file_path_wide))
-
 # Define the path for the wide-format dataset RDS
-file_path_wide_rds <- here("data", "processed-data", "rds", "mosquito_wide.rds")
+file_path_wide_rds <- here("data", "processed-data", "rds", "mosquito_wide_all.rds")
 
@@
 saveRDS(wide_data, file_path_wide_rds)
+
+# Now we will filter the the mosquito data to match the time period of the virus data. 
+# Since mosquito populations can be affected by the previous year's populations, we will filter for 2011-
+
+# Filter wide_data for the specified date range
+filtered_mosquito_data <- wide_data %>%
+  filter(`collection date` >= as.Date("2011-01-01") & `collection date` <= as.Date("2022-12-31"))
+
+# Define the path for the CSV file
+file_path_filtered_csv <- here("data", "processed-data", "mosquito_wide_filtered.csv")
+
+# Save the filtered wide-format data to the specified CSV path
+write_csv(filtered_mosquito_data, file_path_filtered_csv)
+
+# Define the path for the RDS file
+file_path_filtered_rds <- here("data", "processed-data", "rds", "mosquito_wide_filtered.rds")
+
+# Save the filtered wide-format data as an RDS file
+saveRDS(filtered_mosquito_data, file_path_filtered_rds)
 

### 2024-03-15 products/manuscript/Manuscript.qmd
Part 3
Continued work on EDA. Most work for part 3 is contained in the R/eda-code folder. R code will eventually be pulled into the qmd via rds files. Continuing to update the manuscript and add better descriptions to code to facilitate better replication. Still work in progress. All read me files have been updated and will need further attention as the project moves forward.

@@
 
-Eastern equine encephalitis (EEE) is caused by an Alphavirus transmitted
-to humans by the bite of an infected mosquito. Human infections are
-rare, but serious. About 30-50% of symptomatic cases lead to death and
-those who survive are left with life-long disability. The transmission
-cycle for EEE includes two distinct cycles: an amplifying enzootic cycle
-in which the virus is transmitted between the Culiseta melanura mosquito
-and birds and an epizootic cycle in which bridge vector mosquitoes
-transmit the EEE virus (EEEv) from birds to humans and other mammals.
-Because C. melanura feeds almost exclusively on birds, it is not
-considered a direct human threat. However, EEEv isolations in C.
-melanura are an early warning sign that EEEv is circulating in the
-ecosystem. Isolations in bridge vector mosquito species indicate
-heightened transmission risk to humans.
+*This project will inform the development of my dissertation prospectus
+by identifying significant relationships between variables and guide the
+methods development.* 
+Eastern equine encephalitis (EEE) is caused by an
+Alphavirus transmitted to humans by the bite of an infected mosquito.
+Human infections are rare, but serious. About 30-50% of symptomatic
+cases lead to death and those who survive are left with life-long
+disability. The transmission cycle for EEE includes two distinct cycles:
+an amplifying enzootic cycle in which the virus is transmitted between
+the Culiseta melanura mosquito and birds and an epizootic cycle in which
+bridge vector mosquitoes transmit the EEE virus (EEEv) from birds to
+humans and other mammals. Because C. melanura feeds almost exclusively
+on birds, it is not considered a direct human threat. However, EEEv
+isolations in C. melanura are an early warning sign that EEEv is
+circulating in the ecosystem. Isolations in bridge vector mosquito
+species indicate heightened transmission risk to humans.
 
@@
 
-This project will lay the foundation for dissertation research that will
-begin in June, 2024. The research will focus on three
+**This project will lay the foundation for dissertation research that will
+begin in June, 2024.**
 
 # Introduction
-
 ## Background
-
-Humans and other mammals are dead-end hosts for EEE. About 94% of human
-EEE infections are asymptomatic. While symptomatic cases are rare, the
+*Citations will be added when the document is finalized* 
+Humans and other mammals are dead-end hosts for EEE. About 94% of human EEE
+infections are asymptomatic. While symptomatic cases are rare, the
 consequences can be severe. Mortality rates associated with symptomatic
@@
 
-# Methods
-
 ## Study area
-
 Bristol County, Massachusetts covers an area of 691 square miles with a
@@
 Island to the west, Plymouth County, MA to the east, and Norfolk County,
-MA to the north.The county is situated on Buzzard's Bay to the south.
+MA to the north. Naushon Island and Martha's Vineyard are across
+Buzzard's Bay to the Southeast.
 
@@
 (MEL), which is recognized as the primary enzootic vector for Eastern
-Equine Encephalitis (EEE). AWC trees have large roots that form "crypts"
-that hold water throughout the year. These crypts serve as a main
-habitat for MEL larval growth.
+Equine Encephalitis (EEE). AWC trees have large root systems that form
+"crypts" and hold water throughout the year. These crypts serve as a
+main habitat for MEL larval growth and provide shelter during winter
+months.
 
@@
 
-## Data collection and processing
-
-The mosquito surveillance dataset encompasses counts by species from all mosquito traps deployed in Bristol County, MA, from 2007 to 2023.
-The BCMCP, perform mosquito surveillance from June through October annually.
-During this interval, CDC light traps baited with carbon dioxide are placed at various sentinel sites each week.
-Some of the sentinel site have been used for over 50 years.
-The traps are installed in the afternoon and retrieved the following morning.
-Designed to attract mosquitoes in search of blood meals, light traps serve as a reliable measure of the actively feeding mosquito population, thereby establishing them as the standard method for EEE surveillance.
-
-Additionally, the county employs gravid traps, which are specifically aimed at attracting and capturing gravid *Culex* mosquitoes looking for egg-laying sites.
-While these traps are primarily used for West Nile Virus (WNV) surveillance, they hold significant value for EEE monitoring as well.
-Given the role of *Culex* species as bridge vectors for EEE, their capture in gravid traps provides critical insight into the risk of virus transmission.
-
-The  In addition to mosquito
-counts by species, the dataset includes the town or city in which they
-were collected, the collection data, and whether the mosquitoes were
-submitted for virus testing.
-
-While the surveillance data provide insight into mosquito population
-estimates and changes in species composition, it is missing one key
-feature. The virus isolation results are not included in this dataset.
-However, virus isolations in mosquitoes are published on the MDPH
-website by town/city and collection date. With this information, one can
-identify the trap event contained in the dataset which allow for
-calculations of key measure, such as minimum infection rate (MIR).
-
-The surveillance dataset is structured to conform with state and federal
-reporting requirements. As a result, the format of the data must be
-transformed to in order to make it ready for analysis. Once the data
-structure is corrected, it can be joined to other relevant datasets
-using spatial or temporal attributes.
-
-This dataset will be integrated with other datasets using spatial or
-temporal attributes. These datasets will include biological and
-ecological data, environmental and spatial data, and demographic data,
-which are crucial in understanding the transmission dynamics of EEE.
-Potential datasets with their sources are outlined in the table below.
-Other datasets may be included based on availability and need.
-
-| Data                         | Description                                                                                                                                           | Source                                |
-|-------------------|----------------------------------|-------------------|
-| Mosquito Surveillance Data   | Records of mosquito species counts, locations, and dates of collection.
-Data on the presence of EEE in mosquito populations is particularly valuable.
-| XXX County Mosquito Control           |
-| Weather Data                 | Temperature, humidity, rainfall, and wind speed influence mosquito activity, population dynamics, and virus transmission.
-| NOAA                                  |
-| Bird Population Data         | Since birds are a natural reservoir for EEE, information on bird populations and migration patterns could be relevant.
-| eBird                                 |
-| Human Case Reports           | Data on confirmed human cases of EEE, including location, date of onset, and clinical outcomes.
-| MA Dept of Public Health publications |
-| Veterinary Surveillance Data | Since horses and other mammals can also be affected by EEE, veterinary records could provide early warning signs of virus activity in an area.
-| MA Dept of Public Health              |
-
 ## Questions/Hypotheses to be addressed
-
-\*This project will lay the foundation for my dissertation research.
+*This project will lay the foundation for my dissertation research.*
 **The goal for this phase of the project is to answer which
@@
 
-{{< pagebreak >}}
+## Data collection and processing
+The mosquito surveillance dataset encompasses counts by species from all
+mosquito traps deployed in Bristol County, MA, from 2007 to 2023.
+The raw dataset includes over 45,000 records and includes temporal and spatial variable.
+The BCMCP, perform mosquito surveillance from June through October annually.
 
-# Methods
+During this interval, CDC light traps baited with carbon dioxide are
+placed at various sentinel sites each week. Designed to attract
+mosquitoes in search of blood meals, light traps serve as a reliable
+measure of the actively feeding mosquito population, thereby
+establishing them as the standard method for EEE surveillance. Some of
+the sentinel site have been used for over 50 years.
 
-*Specific methods will be determined when all data sources are
-confirmed.*
+Additionally, the county employs gravid traps, which are specifically
+aimed at attracting and capturing gravid *Culex* mosquitoes looking for
+egg-laying sites. While these traps are primarily used for West Nile
+Virus (WNV) surveillance, they hold significant value for EEE monitoring
+as well. Given the role of *Culex* species as bridge vectors for EEE,
+their capture in gravid traps provides critical insight into the risk of
+virus transmission.
 
-## The process will involve creating a relational database in which all datasets are joined using spatial or temporal attributes.
-This will lead to a thorough examination of the datasets to understand the distributions, identify missing data and outliers, and uncover potential correlations between variables.
-The EDA will involve using visualizations, geographic information systems, and statistical tests to understand how biological, ecological, environmental, and demographic forces influence the EEE transmission cycles.
-The EDA will inform the preprocessing steps, feature engineering, and selection of appropriate AI models.
+The traps are installed in the afternoon and retrieved the following
+morning. After the traps are collected, the next are stored in a freezer
+to preserve and kill the mosquitoes. The mosquitoes are sorted by
+species using a dissecting scope. Counts for each species are recorded
+in the database and a portion of the vector species are driven to the
+Massachusetts State Laboratory in Boston. The lab tests each mosquito
+pool (or batch) for WNV and EEEv using PCR. Results are shared with
+BCMCP on the same day of testing. Virus isolation results are recorded
+in the collection records.
 
-### Schematic of workflow
+In addition to mosquito counts by species, the dataset includes the town
+or city in which they were collected, the collection date, and whether
+the mosquitoes were submitted for virus testing.
 
-*This section will be updated before analysis begins*
+One indicator of the level of virus activity present in the mosquito
+population can be calculated from the information contained in dataset.
+The minimum infection rate (MIR) can be calculated at various scales
+from trap site to all sites over an entire season. MIR is calculated as:
+$$
+MIR = \left( \frac{\text{Number of positive mosquito pools}}{\text{Total number of mosquitoes tested}} \right) \times 1000
+$$ 
+MIR provides an estimate of the prevalence of a specific pathogen
+within a mosquito population. While it does not directly measure the
+risk to humans, a higher MIR in mosquitoes that commonly bite humans can
+indicate a higher risk of transmission. MIR allows for the comparison of
+infection rates across different geographical areas or time periods.
+This is crucial for identifying areas or times of higher risk and for
+evaluating the effectiveness of control measures.
+
+![Minimum infection rate by
+week](images/MIR_by_week_heatmap.png){fig-align="center"}
+
+The surveillance dataset is structured to conform with state and federal
+reporting requirements. As a result, the format of the data must be
+transformed to in order to make it ready for analysis. Once the data
+structure is corrected, it can be joined to other relevant datasets
+using spatial or temporal attributes.
+
+Weather data played a significant role in this phase of the analysis.
+Daily weather summaries were downloaded from NOAA's Climate Data Portal.
+The dataset covers the years 2007-2023 at the Taunton Airport in Bristol County.
+
+These datasets will be integrated with other datasets using spatial or
+temporal attributes. These datasets may include biological and
+ecological data, environmental and spatial data, and demographic data,
+which are crucial in understanding the transmission dynamics of EEE.
+Potential datasets with their sources are outlined in the table below.
+Other datasets may be included based on availability and need.
+
+| Data                         | Description                                                                                                                                           | Source                                |
+|-------------------|----------------------------------|-------------------|
+| Mosquito Surveillance Data   | Records of mosquito species counts, locations, and dates of collection.
+Data on the presence of EEE in mosquito populations is particularly valuable.
+| XXX County Mosquito Control           |
+| Weather Data                 | Temperature, humidity, rainfall, and wind speed influence mosquito activity, population dynamics, and virus transmission.
+| NOAA                                  |
+| Bird Population Data         | Since birds are a natural reservoir for EEE, information on bird populations and migration patterns could be relevant.
+| eBird                                 |
+| Human Case Reports           | Data on confirmed human cases of EEE, including location, date of onset, and clinical outcomes.
+| MA Dept of Public Health publications |
+| Veterinary Surveillance Data | Since horses and other mammals can also be affected by EEE, veterinary records could provide early warning signs of virus activity in an area.
+| MA Dept of Public Health              |
 
 ## Data aquisition
-
 With the exception of the mosquito collection data, all other datasets
@@
 ## Data import and cleaning
-
 Datasets will be loaded into R in a Quarto document to document and
@@
 
+*I am deciding how to handle the time period. I do have good data for some variables that span the entire period.*
 The mosquito dataset spans 2007 to 2023. However, the virus and case
@@
 Additionally, the Massachusetts Department of Public Health requires
-that all mosquito control district submit their colelction data in a
+that all mosquito control district submit their collction data in a
 specific format. This format divides each trap event into separate rows
@@
 The bird observation dataset was obtained from Cornell University's
-eBird program. The original dataset included all reported bird
+crowd sourced eBird program. The original dataset included all reported bird
 observations for Southeast Massachusetts and was over 600 MB of data.
@@
 after 2014. The final dataset included 8 of the 50 variables, including
-location data.
+location data. *this dataset is of poor quality. I am looking for a
+better dataset now. Looking into the annual breeding bird survey*
 
-## Statistical analysis
+# Methods
+*This section is under development and will be completed after analysis
+is complete. The following is a general outline of the methods.* All
+datasets with a temporal attribute will be modifided to include a new
+for the week number. Week number calculations will be standardized
+according to ISO 8601 standards and will be stored an a YYYY-WW format.
+For example, week 8 in 2019 will be displayed as 2019-08. Week numbers
+will provide a common field by which all datasets with a temporal
+attribute can be joined. It will also allow for grouping records and
+making calculations by week number. This format helps to standardize
+temporal scales and make comparisons between years easier.
 
-This exercise will serve as the beginning of the exploratory data
-analysis phase. One of the main objectives for of this phase is describe
-the data, understand its structure, clean data, and identify
-associations within the datasets.
+Symptomatic EEE, and to a lesser extent WNV, human cases are rare. On
+average, the United States records only about 9 cases of EEE annually.
+Notably, Massachusetts reports more cases than any other state except
+Florida, highlighting regional disparities in disease occurrence.
+However, the incidence rate for neuroinvasive EEE cases in Bristol
+County, Massachusetts, is estimated to be only around 0.06 per 100,000
+people. The infrequency of these cases poses significant challenges for
+traditional statistical models, making it difficult to accurately
+estimate human transmission risk. This difficulty is largely due to the
+sparse nature of the data and the variability in transmission patterns,
+which complicates efforts to predict outbreaks and necessitates
+alternative modeling approaches.
 
-# Results
+### Schematic of workflow
+*Schematic will be refined in later steps. The doc will also be
+reformatted to move the schematic to an appendix* *The image should
+appear upon rendering*
+
+```{mermaid}
+flowchart TB
+    id1["Questions"] --> id2["Data Collection"]
+    id2 --> id3["Data Loading and Cleaning"]
+    id3 --> id4["Data Cleaning"]
+    id4 --> id5["Data Transformation"]
+    id5 --> id6["Data Exploration & Visualization"]
+    id6 --> id7["Data Modeling"]
+    id7 --> id8["Evaluation"]
+
+    id8 -.-> id1
+    id8 --> id9[Interpretation]
+    id9 --> id10[Communication]
+
+    classDef mainProcess fill:#fff,stroke:#333,stroke-width:2px;
+
+    class id1,id2,id3,id4,id5,id6,id7,id8,id9,id10 mainProcess
+```
 
 ## Exploratory/Descriptive analysis
-
 The first phase of the exploration is complete. The next phase will
@@
 
-The geographic focus for this study is Bristol County. One factor to
-consider is if early virus isolations are associated with cases in
+One factor to consider is if early virus isolations are associated with cases in
 humans and other mammals. The table below shows the date of the first
@@
 ## Basic statistical analysis
+*More will be generated in later phases of the exercise* The initial
+tests have been basic regression models.
 
-*These will be generated in later phases of the exercise*
+The first models explored the relationship between virus isolation in
+mosquitoes and weather conditions (average temperature by week number
+and total precipitation by week). The outcome was whether there was at
+least 1 positive pool in a week. The predictor in the first model was
+the average maximum daily temperature the week before the pool week. The
+logistic model suggest that there was a relationship between the two.
+(The R code for the model can be found in the EDA-Code file: logistic_.pos_temp_precip.R) The average temperature in the week before is positively associated with
+the likelihood of a positive result, with an odds ratio of approximately
+1.069 for each unit increase in temperature (Estimate = 0.06695, SE =
+0.02033, p \< 0.001), indicating that higher temperatures significantly
+increase the probability of a positive outcome. The model's improvement
+is highlighted by a decrease in residual deviance to 405.79 from the
+null deviance of 417.24, upon incorporating Lagged_Avg_TMAX as a
+predictor, demonstrating a better fit to the observed data. However,
+with a residual deviance of 405.79 on 299 degrees of freedom, it is
+evident that not all variability in the outcome is explained, suggesting
+the presence of other influential factors. The model's AIC of 409.79
+indicates a balance between model complexity and predictive power,
+supporting its relative utility in understanding the data.
+However, this relationship should be examined more closely as temperature is likely auto correlated with progressing weeks during summer months.
+While
+precipitation has been demonstrated to influence the population dynamics
+it did not show to have any effect on a positive result.
+
+Another model assessed whether the temperature precipitation in April, May, and June of each year effected the date of the first positive pool test.
+The date of the first positive pool was identified and assigned a day number from January 1st of that year.
+The predictors in this linear model were the average  maximum temperature for April, May, and June and the sum of precipitation for the same time period.
+None of these predictors were significant.
+
+There was one other significant result.
+When MIR is set as the dependent, continuous variable in a linear regression and the average maximum temperature is used as the predictor, the model suggests there is a significant inverse relationship.
+As average temperature increases, the MIR decreases (model_mir_temp.R).
+This relationship is counterintutive, especially given the result of the logitic regression and requires more examination.
+
+More models assessing the effects of weather and other factors on positive pools by mosquito species and MIR values will be developed in Part 4 of the assignment.
+Future models will examine the the effects of weather conditions on mosquito population dynamics.
+
+## Statistical analysis
+This exercise will serve as the beginning of the exploratory data
+analysis phase. One of the main objectives for of this phase is describe
+the data, understand its structure, clean data, and identify
+associations within the datasets.
+
+# Results
+*updated in later phases of the project*
 
@@
 
-*this section will be updated after analysis is complete*
+*this section will be updated after analysis is complete* The accuracy
+and reliability of the mosquito collection data is excellent. The same
+procedures and sentinel sites have were in place for the entire
+analytical horizon. Further, there were only two people responsible for
+sorting, identifying and counting species over the entire period.
 
-While the reliability of the mosquito trap data is high, the trap data
-cannot be directly associated with the virus isolation data. The
-information this scale is not released by the State. Spatial scale is
-also an issue with the human and veterinary cases data. Human case data
-is released only on a county scale to ensure patient privacy. Some
-veterinary case data is available on the town-level, but most remains
-only on the county-level.
+MIR is not a perfect measure of virus activity in mosquitoes. It tends
+to underestimate the level of virus circulating in the environment.
+However, other measures require equal pool numbers for testing. There is
+a correction factor for variable pool numbers and I will explore this
+option in later phases of this project.
 
-These limitations are barriers to accurate spatial analysis.
+The spatial scale is an issue with the human and veterinary cases data.
+Human case data is released only on a county scale to ensure patient
+privacy. Some veterinary case data is available on the town-level, but
+most remains only on the county-level. These limitations are barriers to
+accurate spatial analysis. The project will rely on virus isolation in
+mosquitoes to model risk. The available bird data relies heavily on
+crowd-sourced reports. There is one standardized survey (Breeding Bird
+Survey) that seems to be more accurate and reliable. However, this data
+is not available for every year.
 
@@
 
-References will be added in part 3 of the project
+References will be added in part 4 of the project

### 2024-03-15 R/eda-code/MIR.R

@@ -1,0 +1,160 @@
+library(readxl)
+library(dplyr)
+library(lubridate)
+library(tidyr) # For separate_rows()
+library(here)
+
+# Read the dataset
+pcr_data <- read_excel(here("data", "raw-data", "virus_iso.xlsx"))
+
+# Add week number based on the collection data
+# This code ensures that the ISO 8601 standard is applied to the week number calculation
+# Week 1 of any year is the week that contains January 4th, or equivalently, it's the week that contains the first Thursday of January.
+# Weeks start on Monday and end on Sunday.
+# The last week of the year, week 52 or 53, is the one that contains December 28th.
+# we will use week_num to join tables
+# the format will be YYYY-WW. Week 8 of 2019 will look like: 2019-08
+pcr_data <- pcr_data %>%
+  mutate(
+    # Directly format the collection date into "YYYY-WW"
+    week_num = paste0(year(`Collection Date`), "-", 
+                      sprintf("%02d", isoweek(`Collection Date`)))
+  )
+
+
+# Correctly expand "Both" in Test Type, then adjust to "EEE" and "WNV"
+pcr_data_expanded <- pcr_data %>%
+  mutate(Test_Type_Expanded = case_when(
+    `Test Type` == "Both" ~ "EEE&WNV", # Temporarily mark "Both" for special handling
+    TRUE ~ `Test Type`
+  )) %>%
+  separate_rows(Test_Type_Expanded, sep = "&") %>%
+  mutate(
+    Year = year(`Collection Date`),
+    Week = week(`Collection Date`),
+    Submitted_for_Testing = `Submitted for Testing` == "Y",
+    # Assuming positive results are marked as "Positive" in `Result`
+    Is_Positive = Result == "Positive"
+  )
+
+# Save as RDS
+saveRDS(pcr_data_expanded, here("data", "processed-data", "rds", "pcr_data_expanded.rds"))
+
+# Save as CSV
+write_csv(pcr_data_expanded, here("data", "processed-data", "pcr_data_expanded.csv"))
+
+# Filter records where Submitted for Testing is "Y"
+pcr_data_expanded_filtered <- pcr_data_expanded %>%
+  filter(Submitted_for_Testing)
+
+# Save as RDS
[…]

### 2024-03-15 R/eda-code/Model_mir_temp.R

@@ -1,0 +1,46 @@
+# Load necessary libraries for data manipulation, date handling, and file path management
+library(lubridate)
+library(here)
+library(dplyr)
+
+# Load the datasets:
+# MIR_by_week: Contains information on mosquito pool tests, including the week number and MIR.
+# wx_selected: Weather data with daily maximum temperature (TMAX) and precipitation (PRCP).
+# mosquito_wide: Another dataset related to mosquito data, although it's loaded, it's not used in the subsequent analysis.
+MIR_by_week <- read_csv(here("data", "processed-data", "MIR_by_week.csv"))
+wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))
+mosquito_wide <- read_csv(here("data", "processed-data", "mosquito_wide.csv"))
+
+# Aggregate weather data by week:
+# Calculate the average maximum temperature (Avg_TMAX) and total precipitation (Total_PRCP) for each week.
+# This simplification allows for weekly comparisons with the MIR data.
+wx_aggregated <- wx_selected %>%
+  group_by(week_num) %>%
+  summarise(Avg_TMAX = mean(TMAX, na.rm = TRUE),
+            Total_PRCP = sum(PRCP, na.rm = TRUE),
+            .groups = 'drop')
+
+# Adjust for a one-week lag in weather data:
+# Create a new column (Lagged_week_num) that represents each week's preceding week. 
+# This adjustment accounts for the delayed effect weather might have on MIR.
+wx_aggregated <- wx_aggregated %>%
+  mutate(
+    Lagged_week_num = paste0(
+      if_else(as.integer(substr(week_num, 6, 7)) == 1, as.integer(substr(week_num, 1, 4)) - 1, as.integer(substr(week_num, 1, 4))),
+      "-",
+      sprintf("%02d", if_else(as.integer(substr(week_num, 6, 7)) == 1, 53, as.integer(substr(week_num, 6, 7)) - 1))
+    )
+  )
+
+# Combine the MIR and lag-adjusted weather data:
+# Merge MIR data with weather data based on the adjusted week number, aligning MIR observations with the prior week's weather conditions.
+combined_data_mir <- MIR_by_week %>%
+  left_join(wx_aggregated, by = c("week_num" = "Lagged_week_num"))
+
+# Fit a linear model to analyze the relationship between MIR and weather conditions:
+# Model MIR as a function of the average maximum temperature (Avg_TMAX) and total precipitation (Total_PRCP) from the previous week.
[…]

### 2024-03-15 R/eda-code/PT3_edacode.R

@@ -5,41 +5,82 @@ library(dplyr)
 library(skimr)
 library(ggplot2)
 library(tidyverse)
+library(lubridate)
+library(readr)
+library(forecast)
+library(tsibble) # For time series data wrangling
+library(feasts) # For decomposing and visualizing time series
 
-## ---- loaddata --------
-#Path to data. Note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed_data","processeddata.rds")
-#load data
-mydata <- readRDS(data_location)
+pcr <- read_csv(here("data", "processed-data", "pcr_data_expanded_filtered.csv"))
+wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))
+# Assuming wx_selected and pcr_data_expanded_filtered are already loaded
 
-## ---- table1 --------
-summary_df = skimr::skim(mydata)
-print(summary_df)
-# save to file
-summarytable_file = here("results", "summarytable.rds")
-saveRDS(summary_df, file = summarytable_file)
+# Step 2: Calculate the average TMAX for April, May, and June separately
+avg_temp_by_month <- wx_selected %>%
+  filter(month(DATE) %in% c(4, 5, 6)) %>%
+  group_by(Year = year(DATE), Month = month(DATE)) %>%
+  summarise(Avg_TMAX = mean(TMAX, na.rm = TRUE), .groups = 'drop')
 
-## ---- height --------
-p1 <- mydata %>% ggplot(aes(x=Height)) + geom_histogram() 
-plot(p1)
-figure_file = here("results","height_distribution.png")
-ggsave(filename = figure_file, plot=p1) 
+# Pivot the average temperatures into separate columns for each month
+avg_temp_by_month_wide <- avg_temp_by_month %>%
+  pivot_wider(names_from = Month, values_from = Avg_TMAX, names_prefix = "Avg_TMAX_M") %>%
+  # Ensure all years are present even if some data might be missing
+  complete(Year, fill = list(Avg_TMAX_M4 = NA, Avg_TMAX_M5 = NA, Avg_TMAX_M6 = NA))
 
-## ---- weight --------
-p2 <- mydata %>% ggplot(aes(x=Weight)) + geom_histogram() 
-plot(p2)
-figure_file = here("results","weight_distribution.png")
-ggsave(filename = figure_file, plot=p2) 
[…]

### 2024-03-15 R/eda-code/logistic_.pos_temp_precip.R

@@ -1,0 +1,47 @@
+library(here)
+library(dplyr)
+library(lubridate)
+library(readr)
+
+
+# Load the PCR data set which includes information on mosquito pools tested for viruses, 
+# results of those tests, and the date of collection.
+pcr <- read_csv(here("data", "processed-data", "pcr_data_expanded_filtered.csv"))
+
+# Load the weather data set which includes daily maximum temperatures (TMAX) and other weather variables.
+wx_selected <- read_csv(here("data", "processed-data", "wx_selected.csv"))
+
+# Aggregate the weather data by week to calculate the average maximum temperature for each week.
+# This simplifies the daily temperature data into a more manageable weekly summary,
+# making it easier to correlate with weekly mosquito test results.
+wx_selected <- wx_selected %>%
+  group_by(week_num) %>%
+  summarise(Avg_TMAX = mean(TMAX, na.rm = TRUE)) %>%
+  ungroup()
+
+# Create a lagged version of the average weekly maximum temperature to explore the potential delayed effect of temperature on mosquito test results.
+# The lag function shifts the temperature data by one week, assuming that the impact of temperature on virus transmission might not be immediate.
+weather_weekly <- wx_selected %>%
+  arrange(week_num) %>%
+  mutate(Lagged_Avg_TMAX = lag(Avg_TMAX, 1))
+
+# Aggregate the PCR data by week to determine if there was at least one positive result in each week.
+# This transformation reduces the detailed test results into a weekly summary,
+# indicating the presence or absence of positive tests each week.
+pcr_weekly <- pcr %>%
+  group_by(week_num) %>%
+  summarise(Positive_Result = max(as.integer(Is_Positive), na.rm = TRUE))
+
+# Merge the aggregated PCR data with the lagged weather data by matching their week numbers.
+# This combined data set enables the analysis of the relationship between weekly weather conditions and mosquito test results.
+combined_wx_pcr <- merge(pcr_weekly, weather_weekly, by = "week_num")
+
+# Fit a logistic regression model to examine the influence of lagged average weekly maximum temperature on the likelihood of observing at least one positive mosquito test result in a week.
+# The binomial family specifies that the dependent variable (Positive_Result) is binary,
+# making logistic regression an appropriate modeling choice for this binary outcome.
+logistic_model_temp <- glm(Positive_Result ~ Lagged_Avg_TMAX, family = binomial, data = combined_wx_pcr)
+
[…]
